---
title: 'Carbon and health co-benefits of cycling uptake: methodology (V.1)'
author: "Robin Lovelace"
output: pdf_document
bibliography: ~/Documents/Transport.bib
---

# Introduction

To estimate the relative health/carbon benefits of different scenarios of
behavioural change in the transport sector we use a multilevel model
based on the National Travel Survey (NTS) which provides information at
the regional, household, trip and stage level. The method could be used to
assess the impacts of a wide range of future scenarios. In this paper
we focus on the impacts of cycling uptake under a range of assumptions
about the types of people who are most likely to start cycling.

The method builds on @Crawford2014 but instead of predetermining model outcomes
by fixing the final rate of cycling, each different scenario is based on
different parameters determining *distance decay*. @Crawford2014 used a very
simple formula for distance decay and did not optimise the parameters
$\alpha$ and $\beta$, which was taken from a previous paper [@Iacono2010].

Reproducible code to run the model, in its simplest form, is available
from the 'pct' (short for 'propensity to cycle tool') MIT-licensed GitHub
code/data repository: https://github.com/Robinlovelace/pct. The subsequent
description of methodology is based largely on the
[pswitch-model-NTS.R](https://github.com/Robinlovelace/pct/blob/master/microsimulation/pswitch-model-NTS.R)  script file in this 'repo'.

# Dependencies and data

The model is implemented entirely in open source software based the R language
for statistical analysis. To augment the capabilities of R's 'base' functions,
a number of add-on packages, available from the Comprehensive R Archive Network
([CRAN](http://cran.r-project.org/)) are used:

```{r, eval=FALSE}
pkgs <- c("foreign", "dplyr", "birk", "ggplot2")
```

In the first instance, the model is based completely on the UK's National Travel
Survey (NTS). The latest release (summer 2014) contains data from 2002 to 2012
and is divided into 9 files, 4 of which are used in this study. SPSS
(`*.sav`) files used to load the data into R:

- household.sav (36 MB): household-level data linked via the variable `HouseholdID`.
This table contains data on the local environment as well as household characteristics.
- individual.sav (141 MB): individual-level, with information about individual travellers
and additudes.
- psu.sav (0.16 MB): this contains information about the *Primary Sampling Units* (PSU)
used to ensure a geographically representative stratified sample, including the region
and sub-region of the households within.
- trip.sav (466 MB) the largest file in the 2002-2012 NTS, this file contains
more than 3 million observations of trip-level data with associated
characteristics.

Other tables of interest include stages.sav (which divides multi-stage trips
into shorter steps) and vehicle.sav (information about the vehicle of travel).
These could be used for future refinements of the method. The data were
loaded into R using commands such as the following:

```{r, eval=FALSE}
trips <- read.spss("trip.sav")
```

## Joining the datasets together

In order to select trips by household-level attributes (for example),
the different levels of data must be joined together. This was acheived
using the `*_join` functions from the **dplyr** package. To join individual
with house-level data, for example, the following command was used:

```{r, eval=FALSE}
all <- inner_join(houses, psuid, by = "PSUID")
```

## Sampling

In some cases it will be beneficial to sample from the available data
rather than using the entirety of the 2002 - 2012 dataset which,
as we have seen, is rather large. To sample by sub-region, for example,
the following code can be used:

```{r, eval=FALSE}
psuid <- filter(psuid, PSUStatsReg_B01ID == "Yorkshire/Humberside, Metropolitan")
```

# Allocating probabilities of switching mode

To determine the probability of trips switching to cycling,
a probability was assigned to every trip. We used the `log_cubic()`
function (described in the `pctpack` R package) to translate
distances of trips to probabilities. To enable socio-demographic
and other variables to be accounted for in distance decay,
we divided-up the probability-assignment process into groups.
To illustrate the method, the code below shows how different
parameters into the distance-decay function were used for men and women
(in a full model more than two groups will be used):

```{r, eval=FALSE}
# Load dd parameters from NTS (Anna Goodman)
par_male_urb <- c(0.0001945, -0.002273, -0.1489583, -2.923221)
par_female_urb <- c(-0.0017932, 0.0655261, -0.8396819, -2.892149)

# Probability of urban males switching, based on distance decay
tsam$pswitch <- 0 # create probability variable
tsam$pswitch[tsam$Sex_B01ID == "Male"] <-
  log_cubic(tsam$JD[tsam$Sex_B01ID == "Male"], par_male_urb)

# Probability of urban females switching
tsam$pswitch[tsam$Sex_B01ID == "Female"] <-
  log_cubic(tsam$JD[tsam$Sex_B01ID == "Female"], par_female_urb)
```

Note that any pre-defined distance-decay functions, based on any future
scenarios and divided into many different combinations of groups, could be used here.

## Excluding trips that are impossible to switch to cycling

Some trips simply cannot be cycled. To define these, we identify those
trips (e.g. those already made by bicycle or those made by disabled people)
and set the probability of a switch to zero. The subsequent line
of code, for example, makes impossible for trips made by those with a disability to
switch to cycling:

```{r, eval=FALSE}
tsam$pswitch[tsam$difficulty_travel == "Yes"] <- 0
```
# Generating new cyclists

The model is probabilistic and the hard word goes into estimating
the probability of any given trip switching to cycling based on
distance, mode of travel, socio-demographic characteristics and
other variables. In any case, we will be left with a vector of
probabilities for each scenario: `pswitch`.

To generate new bicycle trips from `pswitch` we use a random number
generator (RNG) to create a new binary variable, `now_cycle`.
A value of 1 means the trip has switched to cycling; a value of
0 means it has not:

```{r, eval=FALSE}
set.seed(666) # ensure results are reproducible
random_num <- runif(n = nrow(tsam), min = 0, max = 1) # random number
tsam$now_cycle <- tsam$pswitch > random_num # new cycle trips
```

# Analysis

Based on the above steps, a wide range of analytic possibilities are opened-up.
Below we simply highlight the potential by creating histograms of number of
trips by distance for trips that switched to cycling and all other trips
(Figure 1). The code below also shows how many new cycle trips are created
compared with the pre-existing modal split.

```{r, echo=FALSE}
setwd("~/repos/pct")
tsam <- readRDS("pct-data/tsam.Rds")
library(ggplot2)
```


```{r, fig.cap="Histograms of trip numbers overall (left) and those which switched to cycling (right) by mode and distance."}
sum(tsam$mode == "Bicycle") # original number of cyclists
sum(tsam$now_cycle) # number of additional cyclists

# This is just and example: see other R scripts for more on analysis
# Distance of new vs existing bicycle trips
ggplot(tsam) +
  geom_histogram(aes(JD, fill = mode)) +
  xlim(c(NA,20)) + xlab("Distance (miles)") +
  facet_wrap(~ now_cycle , scales = "free")
```





# This is just and example: see other R scripts for more on analysis
# Distance of new vs existing bicycle trips
ggplot(tsam) +
  geom_histogram(aes(JD, fill = mode)) +
  xlim(c(NA,20)) +
  facet_wrap(~ ifelse(now_cycle, "Now cycled", "Unchanged") , scales = "free")



# References