---
title: 'Distance decay in active travel: an exploration of functional forms
  for modelling individual and aggregate level data'
author: "Robin Lovelace"
output: pdf_document
bibliography: ~/Documents/Transport.bib
---

Short-distance trip tend to be more frequent than longer ones.
*Distance decay* (henceforth *dd*) is one of a number of concepts that 
neatly expresses the this near-universal observation.
Researchers using the *dd* terminology tend to go beyond verbal
descriptions of the relationship between distance the frequency of trips.
*dd* is a useful term for describing precisely *how* the proportion of trips
($p$) declines with distance ($p$), beyond the general observation that the
slope is negative:

$$
\frac{\delta d}{\delta p} < 0
$$

the term 
has been used to ask, *how* does trip frequency vary as a function
of distance [@Iacono2010, ]. In other words, what is the function for distance decay
($fdd$)?:

$$
fdd(d, X) = tr + e
$$

where $d$ is distance (km), $X$ is a set of explanatory variables, $tr$
is the trip rate, expressed either in absolute numbers or as a proportion of
all trips for a given trip distance and $e$ is stochastic error.
Clearly this equation can be refined. A logical refinement of
is to add subscripts to the terms in equation 1, to
disaggregate distance decay by variables operating at various levels.
Trip type (e.g. trip mode,
and purpose), the characteristics of the people making the trip
(e.g. age and sex) and the location and physical surroundings of the
trip (e.g. hilliness, location) have all been found to affent the shap
of distance decay curves [@Fotheringham1981; @Iacono2010; @Fingleton2005].

Other invocations of the distance decay concept have been developed and it is
worth reviewing these briefly: a variety of words can be used to express
the fundamental truth t, which are axiomatic
in the travel behaviour literature,
include the 'first law of geography', the more vernacular 'friction of distance'
and 'the gravity law'.

Tobler's famous first first law of geography states that
"everything is related to everything
else, but near things are more related than distant things" [@Tobler1970].
The phrase strongly implies that space and distance are important
considerations for understanding the world but does not specify why. Clearly,
the increased frequency of communications and transport between places that
are close to each other can help explain spatial autocorrelation
and related quantitative measures of Tobler's law [@Miller2004a]. However,
the 'first law' is primarily descriptive and says little about *how* the
nature of transport systems contributes to the generalisability of the
'law', or to which proximity is related to the degree of interaction between
two people, places or phenomena. Moreover, in a world of accelerating globalisation
under the auspices of the ongoing 'digital revolution', the relevance of the
'first law' to the system-level processes it was proposed to explain has come under
scrutiny [@Westlund2013].

Limiting the scope of investigation to transportation --- the physical movement
of people and goods, as opposed to communication which can be immaterial
--- can simplify the problem and make it more tangible.
Due to fundament physical limits on the efficiency with which matter
can be moved [@MacKay2009] and a limited supply of energy
(especially pertinent in an era of resource over-extraction and
climate change), transport will always be limited to some degree by distance.
This is well encapsulated in the more theoretical phrase
'the friction of distance' [@Ellegard2004].

The 'gravity model' of movement patterns formalised,
helped to quantify and generalised these ideas [@zipf1946p1].
The gravity model, or 'law' as it has sometimes been called [@]  has been a rich source of theoretical and methodological advance in many fields, primarily
urban modelling but also in fields as diverse as ... and spatial epidemiology
[@Balcan2009].


Despite dissenting voices --- including the statement that "a strict gravity
model simply did not work" for modelling urban systems and that some subsequent
refinements to the gravity model were "fudge factors" [@Wilson1998-past] ---
the gravity model has been one of the dominant tools for understanding
urban mobility over the past 100 years [@masucci2013gravity]. A recent
development in this field has been the 'radiation model'
[@Simini2012]. This new formula for estimating flow rates between geographic
zones is interesting in its ommission of distance as an explicit explanatory
variable, instead using the number 'intervening opportunities' ($IO$)
as a proxy for $dd$ the denominator to estimate flow:

$$
dd \approx (m_i+s_{ij})(m_i+n_j+s_{ij})
$$





a concept that is familiar to conventional transportation planning practice as an
integral part of methods to
model the distribution of trips 

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
