---
title: Where to build bike paths? A generalisable flow-level model of cycling uptake
  for sustainable transport planning
author: "Robin Lovelace"
date: "April 12, 2015"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  word_document: default
bibliography: ~/Documents/Transport.bib
---

```{r, echo=FALSE, include=FALSE}
pkgs <- c("stplanr", "knitr", "xtable", "sp")
vapply(pkgs, require, character.only = T, FUN.VALUE = logical(1))
```


# Introduction

# Data

The basic model requires only two sets of input data
(although more refined versions benefit from many additional datasets):

- *Flow data* estimating the rate of movement between different places,
often recorded through Census travel to work questions. The data is
generally provided as a 'flow matrix' (with rows representing origins
and columns destinations) or a longer table of origin-destination pairs.
This type of data is increasingly available from so-called 'Big Data'
sources such as passively collected mobile telephone records.
- *Geographical data* on the coordinates of trip origins and destinations.
As a bare minimum, this means the (preferably population-weighted)
centroids of each zone in the study area but could also include.

The flow model described in this paper can work for
anywhere that has access to these two data types. To link
the two datasets together, *zone ids* are needed in both
datasets. To ease the process of combining *flow* and *point* data,
an R function called `gFlow2Lines()` was 
created.^[The source code of `gFlow2Lines()` has been made available
online, as part of an R package for sustainable transport planning,
**stplanr**. See [github.com/Robinlovelace/stplanr/](https://github.com/Robinlovelace/stplanr/blob/master/R/gFlow.R).]
The two input datasets and single output, a set of geographically defined
lines with attributes for each flow in both directions
(which we label `flowlines`) are illustrated in the tables (1 and 2)
and figure (1) below.

```{r tflow, echo=FALSE, results='asis'}
t1 <- xtable(head(flow[c(1:3, 12)]), caption = "Sample of the 'flow' input dataset, representing the number of people who commute from locations within and between administrative zones (MSOAs)", label = "tbl:flow")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r tcents, echo=FALSE, results='asis'}
t2 <- xtable(as.data.frame(cents[1:3,-c(3,4)]), caption = "Sample of the 'cents' input dataset, representing the geographical location of the population-weighted centroids of MSOA zones described in Table~\\ref{tbl:flow}", label = "tbl:cents")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r, echo=FALSE, fig.cap="Illustration of 'flow data' converted into geographical lines between origin and destination pairs. Width represents the total number of trips."}
data("flowlines")
plot(flowlines, lwd = flowlines$All / 10)
```


Although the details may differ (e.g. the categories used to disaggregate flow
by vehicle mode, trip type and socio-demographic group), the basic structure
of flow data will be the same the world over.
The model for England described in this paper used the following open datasets:

- `wu03ew_v2.csv`, a 104 MB (12 MB compressed) comma-delimited
(human readable text) file
of flows between unique origin destination pairs, disaggregated by mode (see Table \ref{tbl:flow}). Note that this is a square table, usually loaded as a `data.frame`
in R. Note the origin and destination codes in the top row of table \ref{tbl:flow}
are the same, indicating *intra-zone* flow.
- `cents.geojson`, representing populations-weighted centroids of local
administrative zones (MSOAs). These are loaded in the model from
a small (1.7 MB) human-readable text file stored in the open
data format GeoJSON. This file was originally downloaded as a Shapefile
under the UK's Open Government Licence.^[The license was accessed from http://www.nationalarchives.gov.uk/doc/open-government-licence/version/2/ in
March 2015.]

## Loading data for local government areas

To ensure reproducibility and enable deployment of the model outside the
original case study area of Manchester, a systematic data loading method was
developed. The computational work to load the various datasets is contained
in the folder `loading-data` in the project's central repository.^[See https://github.com/npct/pct/tree/master/loading-data]. 
In the UK, local transport decisions (such as the location of new cycle paths)
are often made at the Local Authoritority level [@Gaffron2003].
 some work to debug the 
data and ensure cross-compatibility of the results) 

# Model scenarios

Once the input data has been processesed and sub-setted to the area of interest,
it is passed to a regression model that estimates what the rate of cycling
*should* be between OD pairs (and hence within local areas via aggregration) based
on a series of explanatory variables. In roughly descending order of importance
(this will be context-dependent) and model implementation these are:

- Distance between origins and destinations. This was calculated in the first
instance as Euclidean (straight line) distance and subsequently refined to an
estimate of route distance using the CycleStreets.net API.

- Hilliness

- Gender balance. The remit for including this variable is that the male:female
ratio of cycle commuters is around 4:1 in the UK; thus including the
proportion of short (5km and less) trips made by females can
greatly improve model fit. Note that this variable was used at the areal
unit level (per zone, as opposed to per origin-destination pair) due to
data limitations. Critically, this variable can be used as the basis of 
scenarios estimating the impact of a 'gender equality' scenario for cycling
increase at the local level.

- Age. 

Note that *total number of trips* is not an explanatory variable. This is because
the dependent variable is not the absolute number of cyclists, but a proportion
of trips made that are expected to be made by bike.

## Baseline

## Gender equality

## Go Dutch

## Electric bicycles

# Results

# Discussion

The flexibility of the approach outlined in this paper ensures that the
method can be used in many different contexts. Because the underlying
methods and computation are transparent and open source, it is
possible to harness them in many different ways. This flexibility
has been demonstrated by the tool's ability to be deployed in any
Local Authority in England, for a range of cycling scenarios, with
the results presented in a range of ways. However, the flexibility
of the approach would allow it to be used in contexts that go well
beyond re-running the model for different cities in the same country.
Potential extensions of the model include:

- Deployment of the tool and underlying flow model in different
countries.^[This would depend on having ]

# Appendix: Using the tool

This Appendix summarises the work from the perspective of practitioners
by describing how the NPCT may be used to inform the decision-making process.

# References
