---
title: Where to build bike paths? A generalisable flow-level model of cycling uptake
  for sustainable transport planning
author: "Robin Lovelace"
date: "April 12, 2015"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
header-includes: \usepackage{amsmath}
bibliography: Transport.bib
---

```{r, echo=FALSE, include=FALSE}
source("../set-up.R")
pkgs <- c("stplanr", "knitr", "xtable", "sp", "grid", "png")
vapply(pkgs, require, character.only = T, FUN.VALUE = logical(1))
```


# Introduction

The National Propensity to Cycle Tool (NPCT) is 
an interactive policy support
application for 
developing transport policies, funded by the UK's Department
for Transport (DfT). The NPCT will be deployed
nationwide and used by Local Authority transport planners
to prioritise investment in cycling. The tool's main purpose
is to help prioritise where to build new strategic cycle paths
and related infrastructure at the city level.

This paper describes the model underlying the
NPCT and explains some of the decisions that were made relating to
the data and methods that it uses.
The model is open source and transparent,
based on input datasets that
are widely available in many countries around the world.
The model is implemented in R, a mature language for
statistical computing. All of the code has been made available
under the MIT license on the GitHub platform.^[See [github.com/npct/pct](https://github.com/npct/pct).]

Although the original model was developed for England, it has the potential to assist with the design of transport policy in many settings.
There is much momentum behind active travel in cities across the world.
Planning support tools such as the NPCT will help the
associated investment to be spent effectively.

```{r, echo=FALSE}
# This transition is needed to prevent runaway
# climate change . The worldwide uptake of cycling, driven by well-placed
# infrastructure,
# 'soft' policy incentives and wider cultural shifts away from the car will also
# have enormous health benefits.
# The design of such infrastructure and policies
# can be enhanced by tools such as the
# NPCT. Given the tool's great potential for wider social benefit, it is
# vital to communicate the underlying methods and software.

# The aim of the present paper is to explain the underlying model which drives
# the scenarios used in the NPCT. This is of academic interest due to
# the increased accuracy with which the method can estimate the
# spatial distribution for cycle infrastructure. It is also of
# great policy relevance during a time of transition away from energy inefficient
# forms of transport such as the personal automobile.
```

The model operates at the flow-level and, by aggregating
data from these flows, at the level of geographic zones. 
A 'flow'in this context constitutes an origin-destination pair: the place where
a trip begins and the destination where it ends. Flows therefore represent
'desire lines', typically connecting residential origins with workplace,
educational or other destinations.
The model takes the current rate of cycling and other variables per flow and uses
this to simulate what the rate of cycling per flow could be
under various scenarios of the future. 
It is important to note that the model is not intended to be predictive and does
not assume the implementation of any specific policies (although further work
could explore adding such capability). 

The model described in this paper
operates 'under the hood' in the NPCT on-line interface.
This means that the
output of the NPCT is simply a visualisation of the model's pre-calculated
outputs. Rather than interacting with the code that drives the model directly,
users interact with visualisations of the model output. This makes the user
interface more accessible. The user is not required to perform any of the model
set-up for the model to run, but can access all of the underlying code
through the project's on-line.^[See
[github.com/npct/pct](https://github.com/npct/pct).]
Thus practictioners can further develop the model if they have the
skills and time to do so, for example to add new explanatory variables
specific to the local area. We plan to encourage a user community to
build around the tool in Phase II of the project, for example,
through advanced training sessions for transport planners with
programming experience.

To communicate the model and the results it produces to
non experts, a 'model output' tab was added.
This feature of the 
user interface, which serves 3 additional purposes: to avoid 'hiding' the
underlying model from users; to encourage modifications
and enhancements to the code; and to ensure transparency.
The tab was also a response to feedback during the user
testing sessions: during user testing
a number of transport
planners had requested further information about the model.
The 'model output' tab
contains key statistics, diagnostic plots and model-results on a per-region
basis. This means that a different summary document is provided depending on
which local authority the user is currently exploring. Some of the outputs of
the Model Output tab are illustrated in Fig. xx to Fig. xx below.
The complete model output document for Manchester is provided in Appendix X.

# Data

The basic model requires only two sets of input data
(although more refined versions benefit from many additional datasets):

- *Flow data* estimating the rate of movement between different places.
Data on mode of travel enables regression models to estimate local
distance decay curves. We use this approach for the NPCT although
it would also be possible to estimate cycling potential in cases where
no break-down by mode is possible.
Flow data is available from various sources including
flows derived from mobile telephone service providers
[@smoreda2013spatiotemporal],
public transport data [@Kitchin2013], household
travel surveys [@TransportforNSW2014] and,
the source used in the model presented below,
Census data on commuting [@Rae2009]. Census data is advantageous due
to its coverage, high geographic resolution and assurances surrounding
data quality. In cases where no census data is available
(e.g. in relation to shopping trips), some combination of the afforementioned
'Big Data' sources
can provide a reasonable approximation of real world travel flows. Flow data is
generally provided as a 'flow matrix' (with rows representing origins
and columns destinations) or a longer table of origin-destination pairs.

- *Geographical data* on the coordinates of trip origins and destinations.
At a bare minimum, this means the centroids of each zone in the study area, preferably weighted by population and/or work location-weighted.  In addition to this, geographic variables could also include additional
features of the urban environment that affect cyclists, such as
hilliness and nodes on the public transport network.

The flow model described in this paper can work for
anywhere that has access to such data. To link
the two datasets together, *zone ids* are needed in both
datasets. To ease the process of combining *flow* and *point* data,
an R function called `gFlow2Lines()` was
created.^[The source code of `gFlow2Lines()` has been made available
on-line, as part of an R package for sustainable transport planning,
**stplanr**. See [github.com/Robinlovelace/stplanr/](https://github.com/Robinlovelace/stplanr/blob/master/R/gFlow.R).]
The two input datasets and single output, a set of geographically defined
lines with attributes for each flow in both directions
(which we label `flowlines`) are illustrated in the tables (1 and 2)
and Fig. x below.

```{r tflow, echo=FALSE, results='asis'}
t1 <- xtable(head(flow[c(1:3, 12)]), caption = "Sample of the 'flow' input dataset, representing the number of people who commute from locations within and between administrative zones (MSOAs)", label = "tbl:flow")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r tcents, echo=FALSE, results='asis'}
t2 <- xtable(as.data.frame(cents[1:3,-c(3,4)]), caption = "Sample of the 'cents' input dataset, representing the geographical location of the population-weighted centroids of MSOA zones described in Table~\\ref{tbl:flow}", label = "tbl:cents")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r, echo=FALSE, fig.cap="Illustration of 'flow data' converted into geographical lines between origin and destination pairs for Coventry. Width represents the total number of trips. Note the use of population-weighted (as opposed to geographic) centroids used for the point of departure."}
# data("flowlines")
# plot(flowlines, lwd = flowlines$All / 10)
grid.raster(readPNG("~/Dropbox/DfT bid/figures/coventry-centroids.png"))
```

Although the details may
differ,^[For example,
categories used to disaggregate flow
by vehicle mode, trip type and socio-demographic group will vary depending on
the data source. Flow data from some sources (e.g. anonymised mobile phone records)
will not have any categories and simply report
total flow.] the basic structure
of flow data is likely to be applicable in many settings.
The model for England described in this paper uses the following open datasets:

- `wu03ew_v2.csv`, a 104 MB (12 MB compressed) comma-delimited
(human readable text) file
of flows between unique origin destination pairs, disaggregated by mode (see Table \ref{tbl:flow}). Note that this is a square table, usually loaded as a `data.frame`
in R. Note the origin and destination codes in the top row of table \ref{tbl:flow}
are the same, indicating *intra-zone* flow.
- `cents.geojson`, representing populations-weighted centroids of local
administrative zones (MSOAs). These are loaded in the model from
a small (1.7 MB) human-readable text file stored in the open
data format GeoJSON. This file was originally downloaded as a Shapefile
under the UK's Open Government Licence.^[The license was accessed from http://www.nationalarchives.gov.uk/doc/open-government-licence/version/2/ in
March 2015.]

## Loading data for local government areas

To ensure reproducibility and enable deployment of the model outside the
original case study area of Manchester, a systematic data loading method was
developed. The computational work to load the various datasets is
undertaken by scripts contained
in the folder `loading-data` in the project's central repository.^[See https://github.com/npct/pct/tree/master/loading-data].
The majority of these only need to be run once; all open-access
data that was created this way was saved in a separate folder:
[github.com/npct/pct-data](https://github.com/npct/pct-data).

In the UK, local transport decisions (such as the location of new cycle paths)
are often made at level of
County and Unitary Authorities (CUAs), of which there are 152 in
England [@Gaffron2003].
Phase I of the NPCT project focussed instead on smaller administrative units:
Local Authority Districts (LADs), of which there are 324 across England
(see Fig. x). For the case study towns of Manchester and Coventry, this
made no difference: LADs and CUAs have the same boundaries for these areas.
For Norwich, however, the LAD is much smaller than the CUA and is less
practical for strategic transport planning at the MSOA level. For this reason
we implemented a buffer selection methodology to expand the scope of the
selection, described in the next section. 
After some work to ensure the cross-compatibility of pre-2011 and post-2011
redefinition of CUAs codes, it was found we could implement the model at the
CUA level.

It would also be possible use other geographic regions as the building
blocks of the model.
Travel to Work Zones (TTWs) are 'commuting watersheds' that correspond to
cohesive regions, the centres of which are known employment centres:
a more scientific version of the NPCT could use TTWs as the regions for building
the model; building the model 'in one go' at the national level would also
be a possibility.

For pragmatic reasons, however, we suggest that the NPCT will be of most use
to transport planners if it operates at the level at which funding is allocated.
As mentioned above, this is at the CUA level.
An increasing proportion of transport funding
in England is also being allocated to Local Enterprise Partnerships (LEPs)
and Combined Authorities (CAs), larger (and often overlapping) geographic
regions.
Based on these insights our
recommendation for Phase II is to build the NPCT for every CUA in
the nation and for selected LEPs and CAs where strategic
cycling plans are being planned. These plans could change based on
feedback from the Department for Transport.

```{r cuas-lads, echo=FALSE, fig.cap="Local Authority District (LAD, above) and County and Unitary Authority (CUA, below) levels of transport planning. These are potential regional units for the National Propensity to Cycle Tool"}
grid.raster(readPNG("~/Dropbox/DfT bid/figures/cuas-las.png"))
```

## Variable zone and flowline selection criteria

In the same way that flow-lines are the unit of analysis for the model,
*zones* are the basic unit of visualisation of the
results over geographic space (it becomes overly complex to 
show more than around 50 flow-lines on the map at any one time).
For reasons related to computational and time resources, it
makes sense to take a sample of the flow lines for the model.
The route allocation algorithm, implemented 'in the cloud' by
CycleStreets.net, is the major bottle-neck at present in terms of
computational time. To make the model scalable and flexible,
we set parameters defining which flows would be sampled.
The most straightforward way to reduce the number of flows
whilst retaining the overall travel pattern is to set a minimum
number on the number of people using it for commuting. Because
the distribution of number of commuters per flow is highly
skewed, a high proportion of commuters can be represented by
a relatively small proportion of flow lines. In Manchester,
for example, 15% of flow lines (those used by 30+ commuters)
account for almost 70% of commuters. The number 30 can be changed
to any number that represents a reasonable balance between being
comprehensive on the one hand and fast to save and load on the other.
Setting the maximum Euclidean distance is another way to reduce the number
of flow lines. We used a value of 15 km, based on the assumption that distances
beyond around 20km are not feasible to cycle on a daily basis. 

In contrast to the flow lines, the main issue with sampling zones was that
there seemed to be *too few* of them at the LAD level to gain a comprehensive
understanding of the travel system. We therefore created another parameter,
`buff_dist`, which represents the distance buffer around the LAD in question
to sample additional zones if the total number sampled without a buffer
fell below some threshold. We set this number to 60 for Manchester
so zones outside the long and thin LAD shape would be selected. This
has the additional advantage of preventing breaks in continuity 
(and zero flows) between different regions in the model.

# Model scenarios

Once the input data has been processed and sub-setted to the area of interest,
it is passed to a regression model. The dependent variable is the 'observed'
level of cycling from the flow data (*olc*); the explanatory variables include
the distance between origin and destination and a series of other flow and
zone-level attributes. By aggregating the origins of all flows,
the model also estimates the rate
of cycling among inhabitants. (Aggregating by destination would also allow
an estimate of the number incoming cyclists per work destination.)
In roughly descending order of importance
(this will be context-dependent) and model implementation these are:

- Distance between origins and destinations. This was calculated in the first
instance as Euclidean (straight line) distance and subsequently refined to an
estimate of route distance using the CycleStreets.net API.

- Hilliness

- Gender balance. The remit for including this variable is that the male:female
ratio of cycle commuters is around 3:1 in the
UK;^[The mean proportion of males cycling to work from the 2011 Census
across Local Authorities
in England is 0.77. This corresponds to a ratio of 3.3:1.]
thus including the
proportion of short (5km and less) trips made by females can
greatly improve model fit. Note that this variable was used at the areal
unit level (per zone, as opposed to per origin-destination pair) due to
data limitations. Critically, this variable can be used as the basis of
scenarios estimating the impact of a 'gender equality' scenario for cycling
increase at the local level.

- Age.

Note that *total number of trips* is not an explanatory variable. This is because
the dependent variable is not the absolute number of cyclists, but a proportion
of trips made that are expected to be made by bike.


```{r, echo=FALSE}
# ## Baseline
#
# The baseline scenario (*baseline*)
```


## Cycling Delivery Plan

The Cycling Delivery Plan scenario (*cdp*) is based on the government's
target to double cycling in England, from 0.8 billion stages currently to
1.6 billion stages by 2025 [@DepartmentforTransport2014].
Assuming increases in population offset
the trend towards lower trip
rates,^[Increases
in population would mean the *proportion* of trips by bicycle
would not need to double for the *absolute number* of trips to double.
However, the rate of trips overall (measured in the number of trips
per person per year by any mode for any purpose) is on a long-term
downward trajectory which more than offsets the impact of population
growth. See
[@Crawford2014] for a detailed discussion of this issue.
The authors could not find evidence about how this trend was
accounted for by the DfT in scenarios from the National Transport
Model (NTM) or elsewhere.]
the Cycling Delivery Plan will result in a doubling in the
proportion of trips made by cycling nationwide.
[@DepartmentforTransport2014] provide no discussion or
breakdown in the geographic distribution of potential cycling uptake in
the scenario, however. The *cpd* scenario should help transport planners
identify where new demand for cycling is likely to be greatest.

Based on this policy context, *cpd* seeks to approximate the geographic
distribution of growth in cycling under a scenario of doubling in cycling.
Cycling does not double in all areas. Places that have high average
commute distances and with an already high rate of cycling will see cycling
increase by less than double; areas with a low current rate of cycling but
high potential based on the number of short commute trips will see cycling
more than double. The same logic applies to flows: origin-destination pairs
that are close together but with a currently low rate of cycling tend to
see the greatest increase in cycling under the  *cpd* scenario.

```{r, echo=FALSE}
# Old description of cpd
# The *cdp* scenario also relates to the *distribution* of trips
# in the study area compared with the data used for the model.
# If the distribution of
# all trips in the regression model
# is flatter than the distribution of cycling trips in the study area,
# a doubling in
# propensity will not lead to a doubling in cycling. If, on the other hand,
# the distribution of cycling trips in the study area mirrors high
# volume flows for which cycling is currently rare,  a doubling in propensity
# will more-than-double the rate of cycling. If the study flows are the same
# as the data used to train the model, *cpd* always results in a doubling in
# cycling. These concepts are illustrated in
# Fig. x below.


# ```{r, echo=FALSE, warning=FALSE, fig.cap="The rate of cycling under the cpd scenario for a case study Local Authority (Coventry). OLC is represented in blue, SLC(cdp) is represented by the white lines."}
# l <- readRDS("../pct-data/coventry/l.Rds")
# flow <- l@data
# mod_logsqr <- glm(clc ~ dist + I(dist^0.5), data = flow, weights = All, family = "quasipoisson")
# flow$fitted <- mod_logsqr$fitted.values * flow$All
# coefs <- coef(mod_logsqr)
# ff <- stplanr::dd_logsqrt(x = flow$dist, a = exp(coefs[1]), b1 = coefs[2], b2 = coefs[3])
# # plot(ff, mod_logsqr$fitted.values) # perfect fit - function works
# flow$cdp <- ff * 2 * flow$All
#
# l2 <- rep(l$dist, l$All)
# l3 <- rep(l$dist, l$Bicycle)
# l4 <- rep(flow$dist, flow$cdp)
# # hist(l2)
# p2 <- ggplot() + geom_histogram(aes(l2), binwidth = 0.5) +
#   geom_histogram(aes(l3), binwidth = 0.5, fill = "blue") +
#   geom_histogram(aes(l4), binwidth = 0.5, col = "white", fill = NA, linewidth = 5) +
#   scale_y_log10() + xlab("Distance (km)")
#
# p1 <- ggplot() + geom_histogram(aes(l2), binwidth = 0.5) +
#   geom_histogram(aes(l3), binwidth = 0.5, fill = "blue") +
#   geom_histogram(aes(l4), binwidth = 0.5, col = "white", fill = NA, linewidth = 5) + coord_cartesian(ylim = c(0, 2000)) + xlab("Distance (km)")
#
# library(gridExtra)
# grid.arrange(p1, p2)
# ```
```

At the heart of the *cpd* scenario is a predictive model of the current rate
of cycling as a function of route distance at the national level. The resulting
distance decay curve is then applied locally. The additional number of cyclists
for each flow is calculated by multiplying the proportion expected for the
distance of the flow nationally. The new rate of cycling ($SLC(cdp)$)
is the current rate of cycling plus this model-based estimate. To demonstrate
how the scenario works, Fig. x illustrates the new and old rate of cycling
in Local Authorities with high and low initial rates of cycling and the
associated shifts in the distribution of trips by distance.

## Gender equality

The Gender Equality scenario (*gendereq*) is a relatively simple modification of
an existing scenario. For the purposes of explanation, we will
use the *observed level of cycling* (*OLC*) as the basis of the
scenario. However, it is possible to apply the *gendereq* method to
any scenario, as described towards the end of this section.

While the Baseline and Cycling Delivery Plan scenarios
(*base* and *cpd* respectively) assume that the local
gender split in cycling continues (typically, around 3/4 of cycle commuters
are males, although this varies over geographic space), *gendereq*
assumes that females become as likely to cycle as males are, for a given
origin-destination pair (per *flow*).
A prerequisite is a model-based estimate of the number of male
and female cyclists between origin and destinations for the baseline scenario.
This involves splitting the number of cyclists project by the model,
the *Scenario-based Level of Cycling*, into
male ($SLC_m$) and female ($SLC_f$) components:

$$
SLC = SLC_m + SLC_f
$$

More males cycle than females in every Local Authority in the country
(Figure x). For this reason, the *gendereq* scenario
is based on the assumption that the rate of cycling amongst
females increases to match the rate of cycling amongst males, rather than vice
versa. Under *gendereq* $SLC_m$ remains constant. The challenge is to
find the value of $SLC_f$ such that the proportion
of females cycling becomes equal to the proportion of males cycling
($pfemale(gendereq)_c = pmale_c$).
Note that this is not as simple as $SLC_f = SLC_m$: the gender split
of the overall rate of flow must be taken into
account.^[To
illustrate this point, consider a flow in which there
are currently more female travellers than males.
$SLC_f$ may rise above $SLC_m$ in this case.
] It is the *proportion* of males and females
per flow who cycle that becomes equal, i.e.

$$
pcycle(gendereq)_f = pcycle_m \\
\frac{SLC(gendereq)_f}{tflowf} = \frac{olc_m}{tflow_m}
$$

where $olc_m$ is the observed level of males cycling,
$SLC(gendereq)_f$ is number of females cycling in the gender equality scenario, and
$tflow_m$ and $tflow_f$ are the total numbers of males
and females in the flow respectively.
There is data about the gender make up of cycling flows at zone level $pcycle(z)$,
and we assume that all flows have this same proportion of male cyclists.
This allows us to estimate the number of males cycling as $olc_m = olc * pcycle_m$,
so that
$$
SLC(gendereq)_f = olc * pcycle_m * \left( \frac{tflow_f}{tflow_m} \right)
$$

and therefore the total flow for gender equality $SLC(gendereq)$would be

$$
SLC(gendereq) = olc_m + SLC(gendereq)_f\\
              = olc * pcycle(z)_m * \left(1 + \frac{tflow_f}{tflow_m}\right)
$$

```{r, echo=FALSE, fig.cap="Cycling and the gender balance of cycling in England. The choropleth maps illustrate the spatial distribution of the two variables. The scatter plot illustrates the relationship between the two variables cycle commuting (x axis) against the proportion of commuter cyclists who are male (y axis) for all 326 Local Authorities (including Districts) in the UK."}
grid.raster(readPNG("../figures/las-gender-pcycle.png"))
```

To illustrate how this method works in practice, imagine a flow in which
5 from a total of 50 people commute by bicycle ($tflow = 50; olc = 5$).
60% of the flow is constituted made by males
($ptrips_m = 0.6$) and 75% of commuter
cycling in the zone is by males ($pcycle(z)_m = 0.75$).
The first stage is to break $tflow$ down into male and female constituents,
of 30 and 20, respectively.  Then

$$
SLC(gendereq)_f = 5* 0.75 * \left(1+ \frac{20}{30} \right) \\
                = 6.25
$$

```{r, echo=FALSE}
tflow <- 50
olc <- 5 #  the currect rate of cycling between origin (o) and destination (d)
ptrips_m <- 0.6 # the proportion of all trips o and d by males
pcyclez_m <- 0.75 # the proportion of cycle trips in the zone/region made by males
tflow_m <- tflow * ptrips_m
tflow_f <- tflow * (1 - ptrips_m)
olc_m <- olc * pcyclez_m
olc_f <- olc - olc_m
pmale_c <- olc_m / tflow_m
slc_gendereq_f <- tflow_f * pmale_c
# print(olc_f)
# print(slc_gendereq_f)
```

The increase to 6.25 cyclists from 5 in the *gendereq* scenario represents
an increase
of 25% from the base scenario. To verify that the numbers make sense,
check if `olc_m / tflow_m` is equal to `slc_gendereq_f / tflow_f`.
The answer is 12.5% in both cases. Gender equality in cycling has been reached.
Note that that $SLC(gendereq)$ can be calculated based on any scenario.
In our model, instead of using the conservative *baseline* scenario,
we estimate $SLC(gendereq)$ based on the Cycling Delivery Plan
($cdp$) scenario.

## Go Dutch

## Electric bicycles

# Results

# Discussion

The flexibility of the approach outlined in this paper ensures that the
method can be used in many different contexts. Because the underlying
methods and computation are transparent and open source, it is
possible to harness them in many different ways. This flexibility
has been demonstrated by the tool's ability to be deployed in any
Local Authority in England, for a range of cycling scenarios, with
the results presented in a range of ways. However, the flexibility
of the approach would allow it to be used in contexts that go well
beyond re-running the model for different cities in the same country.
Potential extensions of the model include:

- Deployment of the tool and underlying flow model in different
countries.^[This would depend on having ]

# Using the tool

This Appendix summarises the work from the perspective of practitioners
by describing how the NPCT may be used to inform the decision-making process.

# References

```{r, echo=FALSE, eval=FALSE}
system("bibtool -x documents/flow-model.aux -o ~/Documents/Transport.bib")
```

