---
title: "National-level analysis of cycling potential"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
bibliography: Transport.bib
---

```{r, include=FALSE}
pkgs <- c("grid", "png", "knitr", "tmap", "dplyr", "ggmap", "pander")
lapply(pkgs, library, character.only = T)
opts_knit$set(root.dir = "../")
# source("analysis/la-analysis.R", print.eval = T)
```

```{r, echo=FALSE}
las <- readRDS("pct-bigdata/national/las.Rds")
# source("analysis/las-tables.R")
```

```{r, echo=FALSE}
# The uptake of active travel operates at many levels.
# At the lowest level
# the decision of which mode of travel to use for a given trip,
# and whether or not to travel at all, is a personal decision
# influenced by local factors. Is my bicycle (or other pedal-powered machine)
# in good working order?
# Is my body ready for the journey? Is the weather nice outside?
# Are the roads scary?
# These are the kind of question that people, consciously or sub-consciously, ask themselves in relation to whether or not to make
# a decision that, evidence shows, could enhance their mental and
# physical well-being: to walk or cycle any particular trip. 

# at the ward level (average population of around 6,500 people)
```


# Introduction

The decision of whether to cycle a particular trip is influenced
by processes operating at many levels. 
Over long time-scales, the terrain, roads, traffic and
local culture will influence the number trips cycled
in a city or region. A growing body of evidence exploring the
relative importance of different factors indicates that
transport policies can influence the rate of cycling at many levels
(see appendix xx).
At the national level, for example,
analysis of 2001 census travel to work data
demonstrated that the proportion of people cycling as their main
mode of commuting
can be explained by hilliness, demographics and the quality of the
local road network, for example [@Parkin2008]. 

The subsequent analysis investigates the current rate and future potential
of cycling in England based on 2011 census data on main mode of travel
to work by distance bands (Parkin's work only considered modal split).
The aim is to provide evidence to decision makers tasked with allocating
funding for strategic funding in cycling across England. Based on the best
available evidence of cycling potential, cost effective interventions
can be made at every level. At the local level this can involve the
decision of whether to construct a cycle path along a rural leisure route or
next to a major A Road. This paper focusses on cycling potential at a much
higher level: the potential for growth in commuter cycling per local authority.
As indicacated in the final section, this analysis could be extended to
estimate cycling potential of increased cycling for all trip purposes, including
education, shopping and leisure.

It is worth considering at the outset the strengths and limitations of the
census commute
data source used in this study. The main advantage is that census data is very comprehensive (with
a very high response rate) and has a high degree of geographical resolution.
Commuting is a trip type that has a high degree of inertia due to the
relatively slow rate at which people move house and place of work.
Commuter trips account for 16% trips made and 19% of distance travelled
nationally and make up a much higher proportion of congested rush hour traffic, where
large benefits can accrue from lower car use. 

Overall commuter cycling is strongly correlated with the rate of cycling
for all trip types [@Goodman2013]. However, in some areas (e.g. with a
high unemployment rate or a high rate of cycling for leisure or shopping
trips) the proportion of people cycling to work may not be representative.
This may also be true of commuting trips in 'commuter belt' areas,
where few people may cycle all the way to work but a higher proportion
cycle for leisure, shopping to or from a local train station
(census commute data only reports the *main mode* of travel).  
In addition, census commuting data are biased towards certain groups
who are fit and able to work and therefore do not represent well many of the young,
the old and the marginalised in society. The use of additional 'data layers'
such as data on the 'school commute' [@Singleton2014a] and estimation of
desire lines for key services such as shops and medical centres will help
overcome these issues in future work.

In this context, the distribution of trip lengths associated with a particular
area is a critical, yet under-explored, factor related to cycling potential. 
To illustrate the point, consider that in some 'commuter belts'
(such as Sevenoaks, described below), the majority of commuting trips are
long-distance and by car.  In such areas, it would not be feasible for most
commuters to cycle all the way to work (although a higher proportion may
incorporate some cycling into their commute trips). In some (primarily rural)
areas, distances to other key services such as schools and shops may also be
high, driving 'car dependency'[@Kay2011].

In high-density urban centres, by contrast, a large proportion of
workplaces and other
destinations are likely to be within distances that are easily cycled.
In the UK context, this helps explain the geographical distribution
in growth/declines in the rate
of cycle commuting. At the national level cycling has grown in recent years
(between the 2001 and 2011 census) in
cosmopolitan cities such as Sheffield, Manchester, Bristol,
Leeds and, most famously, London
whilst stagnating at the national level [@Goodman2013].

In this paper we explore the geographical variability of cycling potential
at the national level with respect to commuting trips, and taking into account
commute distances. Estimates of plausible future rates^[By
'rate' we are referring
to the proportion of trips that
are cycled or the proportion of people who regularly cycle to work.]
or numbers cycling can help inform the decision making process, for example
informing the setting of suitable local targets or the relative allocation of
cycling funding between different local authorities.

All else being equal, it is recommended that the bulk of long-term investment
in cycling is made in areas and along travel corridors where many trips
(or stages) are of an easily cyclable distance. In such areas cycling has the
greatest potential to become "the first and natural choice" for urban transport
[@WelshGovernment2014]. As such, our estimates of cycling potential can
complement other considerations and sources of information such as road safety
statistics and an understanding of the benefits of building momentum
'early adopter' areas where the rate of cycling is already growing.

It has been suggested that "towns with a pre-existing culture of cycling
find it easier to get more people  cycling because they do not have to
overcome conflicting social norms" [@Sloman2014]. On the other hand,
the greatest benefits may be in the long-term accrue to
areas where there is currently a low rate of cycling but high potential
based on the number of short-distance trips taking place.
In any case, cycling potential is clearly related to both the current
rate of cycling and the distance distribution of trips in the area.
These variables are the focus of this paper.

The region of analysis is the entirety of England and is based on Census 2011
commuting data. As mentioned, this dataset has the greatest geographical
resolution and greatest coverage (90% + of the working population, with
a high response rate from every authority) of any travel dataset of England.
The cycle commuting captured in the Census captures over a third of all utility
cycling (Goodman 2013), and may have particularly large benefits for congestion
(by reducing reduce rush hour traffic). The analysis does not seek to estimate
the current or potential level of leisure cycling.
The potential of cycling commuting is also central to the wider National Propensity
to Cycle Tool project, of which this analysis is a part.
The units of analysis are the 324 Local Authority Districts in England (LADs),
the smallest zones that provides a breakdown of mode by distance.

The structure of the paper is as follows. First we describe the
current rate of cycle commuting. Next we explore the potential
for additional cycle trips at the national level, both in terms of
the proportion of all commuter trips cycled and in terms of the
absolute number of commuter cyclists.
Finally we explain the methodology that was used and the impact that this
has on the policy implications of the work.

# The current level of cycling

The spatial distribution of of commuter cycling across England
is depicted in Fig. xx1. The dataset shows that there is wide variation
in the level of cycling, which can be split roughly into
3 categories. In 135 LADs (42% of the total number and population)
less than 2% of workers report cycling as their main means of travel
to work. Nearly half (160) of the 324 LADs have an intermediate level
of cycling, between 2 and 5%. 
The remaining 29 LADs have what can be considered a high rate of
cycling by English standards, of 5% of more. 
This means that only around 10% of the population can be considered
as living in areas where cycling is a 'normal' mode of travel to work.
8 of the 'high cycling' LADs are
in London.^[These are: Hammersmith and Fulham, Wandsworth, Hackney,
Tower Hamlets, Islington, Richmond upon Thames,
Southwark and Lambeth.]


```{r, echo=FALSE, fig.cap="The current level of cycling to work in England from the 2011 Census."}
fixedBreaks = c(0,2,5,30)
qtm(las, "pcycle", fill.style = "fixed",
  fill.breaks = fixedBreaks, fill.title = "Percent cycling")
```

Outside London, 'high cycling' LADs tend to be mid-sized high-density urban
areas focussed on a large mono-centric centre. Cambridge (with 29% cycling),
Oxford (17%) and York (11%) are well known high-cycling Local Authorities
that fit this category but the list also contains lesser-known
areas. Norwich (9%), Bristol (7%), Lincoln (6%) and Portsmouth and
Gosport (with 7 and 10% cycling to work, respectively) also have
a relatively high rate of commuter cycling. 
Further details on the top 10 LADs by cycling share in commuting
are provided Table 1. Alongside the observed % cyclists, we have put the
'expected %', as calculated from national average cycling levels
plus the distribution of commute distances in each city.
This shows that the observed proportion of cyclists in the areas shown in Table 1
is around two-to eight-times larger than the background national average.
The final column of Table 1 shows the proportion of commuters who travel
very short distances to work (less than 2km) who drive to work, representing the
extent to which cars dominate the local transport system.
This final column is revealing because it shows firstly that high cycling areas
are less car-dependent than low cycling areas (compare with Table 2) but also that
cars continue to dominate local travel to a
considerable extent even in some relatively high-cycling areas.

Table: The top 10 'high cycling' Local Authorities in terms of the
proportion of commuters who report cycling as their usual main mode
in the 2011 Census ('Observed n cycling').
The 'expected commuter cyclist' figures are calculated based on the
distribution of commute distances in each local authority
and the average national cycle mode share. See
https://github.com/npct/pct-data/blob/master/las_data.csv for the full table.

```{r, echo=FALSE}
t0 <- tail(las@data[order(las$pcycle),], 10)
t0 <- t0[10:1,]

t0 <- select(t0, Name = NAME, `Observed n cycling` = Bicycle_All, `Observed % cycling` = pcycle, `Expected n` = expected, `Expected %` = pcycle_exp, `Car mode share for v. short commutes` = shortcar)
kable(t0, digits = 1, row.names = F)
# pander::pander()
```

Areas with the lowest rate of using cycling as the 'usual main mode' for
commuting are presented in Table 2. It is interesting to note that these are not
exclusively rural areas, where low cycling could be explained by the lack of
jobs near to houses. Five of the 10 Local Authorities with the lowest level
of cycling all the way to work are largely rural: Rossendale, Harrow, Tandridge,
Bolsover and North East Derbyshire.
These are by no means the *most* rural areas in
England and, furthermore, each 
lies in close proximity to motorways and other
large roads.
Bradford seems like an outlier in this group as it is
a city with well-defined commercial centre. However,
the Local Authority also contains large expanses of rural areas,
so can to some extent be grouped with the aforementioned areas.

Two of the areas in the 'bottom 10' for using cycling as the usual main mode
are classic commuter belts: Sevenoaks and Epping Forest are wealthy areas on
the outskirts of London. The remaining two 'low cycling' LADs presented in
Table 2 are relatively deprived urban areas that are part of a wider conurbation:
Barnesley and Rotherham. These areas
have car-dominated and linear urban centres, as opposed to York and
Cambridge which have partially pedestrianised city centres.
The complete dataset, containing information for all LADs,
can be found online at [github.com/npct/pct-data/blob/master/las_data.csv](https://github.com/npct/pct-data/blob/master/las_data.csv).

Table: The bottom 10 Local Authorities in terms of the proportion of
commuters who report cycling as their usual main mode in the 2011 Census
('Observed n cycling').  The 'expected %' figures are calculated based on the
commute distances and the average national cycle mode share.

```{r, echo=FALSE}
t0.1 <- head(las@data[order(las$pcycle),], 10)

t0.1 <- select(t0.1, Name = NAME, `Observed n cycling` = Bicycle_All, `Observed % cycling` = pcycle, `Expected n` = expected, `Expected %` = pcycle_exp, `Car mode share for v. short commutes` = shortcar)

kable(t0.1, digits = 1, row.names = F)
```

Given the wide range of current levels of cycling across England,
it makes sense to allocate funding to each of the 3 types of
Local Authority. Authorities with a high current level of cycling
generally have a strong level of commitment to and community
support for cycling. New infrastructure along key 'cycling desire lines'
(e.g. as identified locally by the NPCT model) would likely
attract the 'near market' for cycling in these areas.
Areas with lower current levels of cycling may require greater
emphasis on complementary behavioural interventions in order to maximise the
value of infrastructure investment [@Sloman2014].

## Short commutes and cycling

We defined ‘short’ commute trips as those below 5km in Euclidian distance.
Based on circuity (or diversion factor) values of 1.3, this corresponds
to a maximum road distance of around 6.5 km.  Trips below 2km Euclidean distance
were defines as ‘very short’. Fig xx2 illustrates the proportion and absolute
number of short commute trips (<5 km euclidean distance) in each LAD.

Because the majority of new cycling trips are short, this data
can be used as a proxy for the potential level of cycle commuting
(although it is recognized that  this proxy is less good in areas where
multi-stage commuting trips are common, e.g. where many people
cycle to bus or train stations before using the public transport as their main mode).

```{r, echo=FALSE, fig.cap="The geographical distribution of short trips in England (< 5 km in Euclidean distance)."}
las$`Short trips (n)` <- las@data$All_.0.2. + las@data$All_.2.5.
las$`Short trips (%)` <- las$`Short trips (n)` / las@data$All_All * 100
qtm(las, fill = c("Short trips (n)", "Short trips (%)"), fill.n = 3, fill.palette = "Blues", fill.style = "jenks")
```

There is a correlation between the proportion of commutes in
an area that are short and the proportion of commutes that are
made by cycle (Fig. xx). Although the relationship is not
particularly strong ($R^2 = 0.17$), the plot is useful for
identifying the areas with a lower than expected rate of commuter cycling
based on current trip distributions. Points below the line in
Fig. xx have disproportionately low level of cycling based on
commuting behaviour, indicating high future potential update.
The LADs with the largest negative residual have a much lower
rate of commuter cycling than would be expected from the proportion of
short trips in the area. The area with the single largest negative
residual is Hastings. The area with the largest positive residual is Cambridge.
These are illustrated in Fig. xx with red dots and accompanying text. The area with the
largest positive residual is Cambridge. These are illustrated
in Fig. xx with text.

The top 10 areas with the greatest unrealised potential for cycling by
this measure are, in descending order,
Hastings, Blackburn with Darwen,
Burnley,                      Chesterfield,
Bradford,                     Oldham,
Hartlepool,                   Torbay,
Stoke-on-Trent  and             Calderdale.
Each of these local authorities are represented by a blue circle in
Fig. xx3.

```{r, echo=FALSE, fig.cap="Scatterplot illustrating the link between the proportion of commutes that are short and the proportion of commutes that are cycled."}
lmod1 <- lm(pcycle ~ `Short trips (%)`, data = las@data)
# summary(lmod1)
sel <- c(which.min(lmod1$residuals), which.max(lmod1$residuals))
sel2 <- head(order(lmod1$residuals), 10)
sel3 <- tail(order(lmod1$residuals), 10)
# las$NAME[sel3]
# library(cowplot)
ggplot(las@data, aes(`Short trips (%)`, pcycle)) + geom_point() +
  geom_abline(intercept = -0.922, slope = 0.1092, col = "red") +
  geom_text(data = las@data[sel,], aes(`Short trips (%)`, pcycle,  label = NAME)) +
  geom_point(data = las@data[sel,], aes(`Short trips (%)`, pcycle), color = "red") +
  geom_point(data = las@data[sel2,], aes(`Short trips (%)`, pcycle), shape = 1, size = 5, col = "blue") +
  ylab("Level of cycling (%)")
  # geom_line(aes(lmod1$fitted.values, `Short trips (%)`), col = "red")
las$short_trip_residual <- lmod1$residuals
# write.csv(las@data, "pct-bigdata/national/la-ttw-distance.csv")
# This clear pattern The relationship between the average *length* of trips and the current ...
```

The analysis suggests that cycling interventions in
such areas (with more than around 40% of trips being less than 5 km
yet a low rate of commuter cycling) could, if effective, yield a high return on investment.
However, we note that the *cycling culture* of these areas is likely
to be weak at present.  As such, it may take time and 'soft' policy interventions (e.g. complementary behavioural interventions), in addition to high quality cycle infrastructure,
to realized the potential for cycling. 

Fig xx3 also illustrates the fact that even those areas furthest above the red
line still consistently have a level of cycling that is far lower than the
proportion of short trips.  For example, the cycle commute mode share of 28.9%
in Cambridge is still under half the proportion of commute trips that are ‘short’ in that local authority (58%).  Similarly, the cycle commute share of 7.7% in South Cambridge is only a third of the 21% of commuting trips that are ‘short’.

Thus even these higher-cycling areas may still have a considerable potential for cycling to grow further. One potential advantage of promoting cycling (e.g. via new high quality cycle paths, cycle parking and disincentives to drive) in an area that already has a strong cycling community is that full use of new infrastructure could be made more quickly than in other areas (Sloman et al. 2014). Thus, new infrastructure along key 'cycling desire lines' (e.g. as identified locally by the NPCT model) would likely attract the 'near market' for cycling in these areas.

# The national distribution of cycling potential

Building on the descriptive analysis presented above, this
section estimates cycling potential under the 'Cycling Delivery Plan'
scenario of cycling uptake, described in Appendix xx.
Applying the flow model to flows with distance distributions
representative of each authority allows for the scenarios described
in section xx to be applied nationwide. The results are presented
in Fig. 3 at the level of County and Unitary Authorities, the level
at which most transport investment is currently made.

Note that 'cycling potential' can be measured in
absolute or proportional terms and that authorities with a higher
population tend to have a higher absolute potential increase in cycling.
The distinction may be important when allocating funding, as the different metrics may be useful for different types of decision making.  Absolute numbers may be advantageous when allocating funding based on estimated economic values of extra cycling activity. Relative measures may be advantageous when designing local targets or assessing whether local areas should meet or exceed national-level targets.

Fig. x illustrates the geographical variability in the potential
increase in the rate of commuter cycling under the Cycling Delivery Plan
Scenario. It is important to note that this is based *only* on the
trip distribution of commuter trips in each local authority;
hilliness is not taken into account. The values generated
correlate well with the observed level of cycling in the 2011
Census: $R^2 = 0.17$. Note that this is roughly the same level of correlation
that was found between the percentage of short trips and the level
of cycling presented above. 

```{r, echo=FALSE, fig.cap="Geographical variability in the potential of cycling to grow across England based on commuter trip distance distributions under the Cycling Delivery Plan scenario, in percentage (left) and absolute (right) terms."}
las$pcycle_cdp <- las$pcycle + las$pcycle_exp
las$Bicycle_exp <- las$All_All * las$pcycle_exp / 100000
fixedBreaks2 <- c(0,2,3,5)
fixedBreaks3 <- c(0,1,2,20)
qtm(las, c("pcycle_exp", "Bicycle_exp"),
  fill.style = c("fixed", "fixed"),
  fill.breaks = list(fixedBreaks2, fixedBreaks3), fill.title = c("Growth in cycling (%)",
    "Growth in cycling\n(1000 commuters)"))
```


Table xx presents the top 10 Local Authorities in terms of increased number of commuter cyclists under our Cycling Delivery Plan.  Aggregated to the level of County and Unitary Authorities (CUAs), the level at which many transport decisions are made, the results are also are presented in Fig. x.  Under our implementation of the Cycling Delivery Plan, the top 10 CUAs in terms of estimated potential growth in absolute numbers of cycle commuters are: Lancashire, Hampshire, Kent, Essex, Birmingham, Hertfordshire, Surrey, Leeds, Staffordshire and Nottinghamshire.
See [github.com/npct/pct-data/blob/master/las_data.csv](https://github.com/npct/pct-data/blob/master/las_data.csv) for the complete dataset.

Table: Top 10 Local Authority Districts (LADs) in terms of potential level of commuter
cycling under the Cycling Delivery Plan (CDP) scenario. Note the raw data
for all LADs can be found online at github.com/npct/pct-data/blob/master/las_data.csv

```{r, echo=FALSE}
las$Bicycle_exp <- round(las$Bicycle_exp * 1000)
t0.1 <- tail(las@data[order(las$Bicycle_exp),], 10)

t0.1 <- select(t0.1, Name = NAME, `Observed n cycling` = Bicycle_All, `Observed % cycling` = pcycle, `Expected n` = expected, `Expected %` = pcycle_exp, `Car mode share for v. short commutes` = shortcar)
t0.1 <- t0.1[10:1,]
kable(t0.1, digits = 1, row.names = F)

# save las data
lasdat_df <- select(las@data, Name = NAME, `N. commuters (all modes)` = All_All, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `Car mode share for v. short commutes` = shortcar, `Short trips (%)`)
# write.csv(lasdat_df, "pct-data/las_data.csv")
```
 
Note that the size of these CUAs plays a strong role in
their potential for increased cyclists in absolute terms.
The 10 CUAs listed above have an average employed population
of more than 500,000. The average number of working people in all
CUAs, by contrast, is 170,000.
(For comparison, the average per LAD is just under 80,000). These numbers underline the importance
of measuring expenditure in pounds per person for areas, rather than
simply estimating the overall total. To see a summary of the results at the CUA level,
please navigate to [github.com/npct/pct-data/blob/master/cuas_dat.csv](https://github.com/npct/pct-data/blob/master/cuas_dat.csv). 

Figure xx illustrates the same information but in percentage terms and is therefore unaffected by the size of the CUA. At the CUA level the the greatest percentage point increases in cycling according to our method (exluding London boroughs) are Hull, Leicester, Plymouth, North East Lincolnshire, Bristol, Nottingham, Derby, Stoke-on-Trent and Newcastle upon Tyne.
These LADs are closely followed by
Middlesbrough, Hartlepool, Portsmouth, Southampton, Blackpool, Liverpool, Poole,
Torbay, Swindon, Manchester and many other towns in terms of estimated cycling
growth potential. See [github.com/npct/pct-data/blob/master/cuas_dat.csv](https://github.com/npct/pct-data/blob/master/cuas_dat.csv) for the complete dataset.

```{r, echo=FALSE, fig.cap="Potential growth in cycle commuters in absolute terms at the level of highway authorities in England. The full dataset can be viewed online: github.com/npct/pct-data/blob/master/cuas_dat.csv", message=FALSE}
# Do for cuas too?
library(rgdal)
cuas <- raster::shapefile("pct-bigdata/national/cuas.shp")
lasp <- rgeos::gCentroid(las, byid = T)
ldf <- las@data
output_data <- select(ldf, Bicycle_exp, All_All)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$Bicycle_exp <- cuasp@data$Bicycle_exp / 1000
qtm(cuas, "Bicycle_exp", fill.title = "Growth in commuter\ncyclists (1000)")
# cuas@data$CTYUA12NM[order(cuas@data$Bicycle_exp, decreasing = T)]
output_data <- select(ldf, All_All)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$All <- cuasp@data$All_All / 1000
cuas$All_All <- cuasp@data$All_All
cuas$Bicycle_exp <- cuas$Bicycle_exp * 1000
```

```{r, echo=FALSE, fig.cap="Potential growth in cycle commuters in percentage points across highway authorities in England. The full dataset can be viewed online: github.com/npct/pct-data/blob/master/cuas_dat.csv", message=FALSE}
cuas$`Growth in commuter cycling (%)` <- cuas@data$Bicycle_exp  / cuas@data$All_All * 100
qtm(cuas, "Growth in commuter cycling (%)", fill.style = "fixed", fill.breaks = c(0,2.5,3,5))
cuasdat <- select(cuas@data, Name = CTYUA12NM, `All commutes (n)` = All_All, `Growth in cycling (n)` = Bicycle_exp, `Growth in commuter cycling (%)`)
# summary(cuasdat)
# summary(head(cuas@data$All[order(cuas@data$Bicycle_exp, decreasing = T)], 10))
# head(cuas@data)
# cuasdat[order(cuasdat$`Growth in commuter cycling (%)`, decreasing = T),]
# sum(cuas@data$Bicycle_exp, na.rm = T) / sum(las@data$Bicycle_All, na.rm = T) * 1000
# write.csv(cuasdat, "pct-data/cuas_dat.csv")
```


# Method and discussion

The results presented in this paper are designed to provide a first
approximation of how the level of cycling that is attainable varies
across different
LADs in England depending on current trip distance distributions.
We are confident that the fundamentals of the
analysis are robust, based on very reliable
census data on the distances that
people travel to work at relatively high levels of geographical disaggregation.

The work underlines the fact that the current distribution of trips means that the cycling potential of some areas is higher than others.  For example, it would be challenging for a commuter belt such as Sevenoaks, where only 19% of commuter trips are below 5 km, to achieve a high rate of cycling *all the way to work*.^[Integrating cycling with public transport in such areas, e.g. by
encouraging cycling to bus and train stations, may allow a higher
proportion of trip stages to be made by bicycle in such areas, however.
In Sevenoaks 52% of commutes are made by car and 20% are made by rail.
Note that in
in many other such
'commuter belts' the current and potential rate of rail use will be lower.
]
Such considerations should inform strategic investments and policies promoting active travel, by highlighting areas where the distribution of current trip distances is most cycle friendly.  Identifying such areas is an important step towards creating a nationally 'joined-up' cycling strategy and allocating funding based on evidence.

There are a number of ways that the analysis could be extended,
including:

- The extension of the method to all trip purposes, for example based on
a version of the National Travel Survey that contains LAD-level data or based
on modelled trip distances.
- Presenting the rate of cycling per 'short trip' instead of per trip
of any length - this would highlight areas that have done well despite
local trip-distance distributions.
- The addition of hilliness per LA as a predictor of the potential level of cycling.
- Using weather-based variables, such as number of rainy days per year, as explanatory variables.
- The use of transport infrastructure variables, such as the proportion
of population living within a certain distance of motorways and A roads, as explanatory variables.

The addition of additional
explanatory variables should improve the predictive
capability of the model. However, it is important to note that the
purpose of the model is not to predict the current rate of commuter cycling,
but to estimate what the plausible rate of cycling could be under
scenarios of the future to guide strategic active travel policies.
An additional extension of this national-level
analysis would therefore be the estimation of cycling potential under
different future scenarios, other than the Cycling Delivery Plan. The
'Gender Equality', 'Go Dutch' and 'Ebikes' scenarios, used in the
flow-level model, would be logical extensions to the analysis in this
direction.

A wider point is that when the flow model described in Appendix xx
is deployed nationwide, model-based outputs per CUA would contain
estimates of the future level of cycling under each of the aforementioned
scenarios. This would minimise duplication in the workflow and integrate
the analysis of the flow data and the national data presented in this paper.

```{r, echo=FALSE}
# A question related to funding is the level at which cycling
# investment will be allocated. If cycling policies are to form
# part of an integrated national transport strategy, it is recommended
# that funding for cycling is allocated at the level at which transport
# investment for other modes takes place. Currently this is the level
# of Combined and Unitary Authorities (CUAs) and, increasingly,
# Combined Authorities and Local Enterprise Partnerships (LEPs).
# CUAs with high unmet potential for cycling include
# Lancashire, Hampshire and Kent. Our analysis suggests that
# funding for cycling should move away from
# sporadic and time-limited city-by-city investment to
# a national-level strategy. Long-term funding for every Highway
# Authority dedicated to cycling will enable a joined-up
# and cost-effective cycling network to be created nationwide.
```

# References

```{r, echo=FALSE, eval=FALSE}
file.copy("documents/national.docx", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.docx", overwrite = T)
file.copy("documents/national.pdf", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.pdf", overwrite = T)
```

