---
title: "National-level analysis of cycling potential"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
bibliography: Transport.bib
---

```{r, include=FALSE}
pkgs <- c("grid", "png", "knitr", "tmap", "dplyr", "ggmap")
lapply(pkgs, library, character.only = T)
opts_knit$set(root.dir = "../")
# source("analysis/la-analysis.R", print.eval = T)
```

```{r, echo=FALSE}
las <- readRDS("pct-bigdata/national/las.Rds")
# source("analysis/las-tables.R")
```

```{r, echo=FALSE}
# The uptake of active travel operates at many levels.
# At the lowest level
# the decision of which mode of travel to use for a given trip,
# and whether or not to travel at all, is a personal decision
# influenced by local factors. Is my bicycle (or other pedal-powered machine)
# in good working order?
# Is my body ready for the journey? Is the weather nice outside?
# Are the roads scary?
# These are the kind of question that people, consciously or sub-consciously, ask themselves in relation to whether or not to make
# a decision that, evidence shows, could enhance their mental and
# physical well-being: to walk or cycle any particular trip. 
```


# Introduction

The decision of whether to cycle a particular trip is influenced
by processes operating at many levels. 
Over long time-scales, the terrain, roads, traffic and
local culture will influence the number trips cycled
in a city or region.
Regression analysis of geographical data at the ward level,
with an average population of around 6,500 people,
has demonstrated that the rate of cycling to work
can be explained by some of these variables
[@Parkin2008]. 

In this context, the distribution of trip lengths associated with
a particular area is a critical, yet under-explored, factor related to
cycling potential.
In some 'commuter belts' (such as Sevenoaks, described below),
for example, the majority of trips are
long-distance. This precludes a strong utility cycling culture without
substantial changes in travel demand.
In high-density urban centres, by contrast, a large proportion of
travel is within a distance that is easily cycled.
In the UK context, this helps explain why the rate
of cycling has grown rapidly
cosmopolitan cities such as Sheffield, Manchester, Bristol,
Leeds and, most famously, London
whilst stagnating at the national level [@Goodman2013].

In this paper we explore the geographical variability of
*cycling potential* at the national level.
Estimates of the future rate^[By
'rate' we are referring
to the proportion of trips that
are cycled or the proportion of people who regularly cycle to work.]
or numbers cycling can be critical for strategic decision-making

The former can provide an
evidence base to inform the setting of suitable local
cycling targets; the latter can assist with the relative allocation of
funding for cycling investment between different local authorities.
Both can help inform the decision making process,
alongside other considerations and sources of information
such as road safety statistics and an understanding of the benefits of
building on the moment behind cycling in 'early adopter' areas:
it has been suggested that "towns with a pre-existing culture of cycling
find it easier to get more people  cycling because they do not have to overcome conflicting social norms" [@Sloman2014].
Nevertheless, the long-term potential is clearly related to the current
rate of cycling. This is the focus of this paper.

The region of analysis is the entirety of England and,
although now slightly out of date, is based on commuting data.
This has the greatest coverage (90% + of the working population),
geographical resolution and coverage (with a high response rate from
every authority) of any travel dataset of England. Commuting data
also reflect utility cycling, as opposed to leisure cycling, which is
the focus of the National Propensity to Cycle Tool project, of which
this analysis is a part.
The main unit of analysis is the 324 Local Authority Districts in England,
the highest resolution data that provides a breakdown of mode by distance.

The structure of the paper is as follows. First we describe the
current rate of cycling. Next we explore the potential
for additional cycle trips at the national level, both in terms of
the proportion of all commuter trips cycled and in terms of the
absolute number of commuter cyclists.
Finally we explain the methodology that was used and the impact that this
has on the policy implications of the work.

# The current level of cycling

The spatial distribution of of commuter cycling across England
is depicted in Fig. xx1. The dataset shows that there is wide variation
in the level of cycling, which can be split roughly into
3 categories. In 135 LADs (42% of the total number and population)
less than 2% of workers report cycling as their main means of travel
to work. Nearly half (160) of the 324 LADs have an intermediate level
of cycling, between 2 and 5%. 
The remaining 29 LADs have what can be considered a high rate of
cycling by English standards, of 5% of more. 
This means that only around 10% of the population can be considered
as living in areas where cycling is a 'normal' mode of travel to work.
8 of the 'high cycling' LADs are
in London.^[These are: Hammersmith and Fulham, Wandsworth, Hackney,
Tower Hamlets, Islington, Richmond upon Thames,
Southwark and Lambeth.]


```{r, echo=FALSE, fig.cap="The current level of cycling to work in England from the 2011 Census."}
fixedBreaks = c(0,2,5,30)
qtm(las, "pcycle", fill.style = "fixed",
  fill.breaks = fixedBreaks, fill.title = "Percent cycling")
```

Outside London, 'high cycling' LADs tend to be high density urban
areas focussed on a large mono-centric centre. Cambridge (with 29% cycling),
Oxford (17%) and York (11%) are well known high-cycling Local Authorities
that fit this category but the list also contains lesser-known
areas. Norwich (9%), Bristol (7%), Lincoln (6%) and Portsmouth and
Gosport (with 7 and 10% cycling to work, respectively) also have
a relatively high rate of cycling. 
Further details on the top 10 LADs by cycling share in commuting
are provided Table 1.

Table: The top 10 'high cycling' Local Authorities in terms of the
proportion of commutes made by bicycle from the 2011 Census.

```{r, echo=FALSE}
t0 <- tail(las@data[order(las$pcycle),], 10)
t0 <- t0[10:1,]

t0 <- select(t0, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)
kable(t0, digits = 1, row.names = F)
```

Areas with the lowest cycling rate are presented in Table 2.
It is interesting
to note that these are not exclusively rural areas, where low cycling could
be explained by the lack of jobs near to houses.
Five of the 10 Local Authorities with the lowest level of cycling to work
are largely rural: Rossendale, Harrow, Tandridge,
Bolsover and North East Derbyshire.
These are by no means the *most* rural areas in
England and, furthermore, each 
lies in close proximity to motorways and other
large roads.
Bradford seems like an outlier in this group as it is
a city with well-defined commercial centre. However,
the Local Authority also contains large expanses of rural areas,
so can to some extent be grouped with the aforementioned areas.

Two of the areas in the 'bottom 10' for commuter cycling are
commuter belts: Sevenoaks and Epping Forest are wealthy areas
on the outskirts of London. 
The remaining two 'low cycling' LADs presented in Table 2 are
relatively deprived urban areas and which are part of a
wider conurbation: Barnesley and Rotherham. These areas
have 'run-down' centres that are car-dominated.

Table: The bottom 10 Local Authorities in terms of the
proportion of commutes made by bicycle from the 2011 Census.

```{r, echo=FALSE}
t0.1 <- head(las@data[order(las$pcycle),], 10)

t0.1 <- select(t0.1, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)

kable(t0.1, digits = 2, row.names = F)
```

Given the wide range of current levels of cycling across England,
it makes sense to allocate funding to each of the 3 types of
Local Authority. Authorities with a high current level of cycling
generally have a strong level of commitment to and community
support for cycling. New infrastructure along key 'cycling desire lines'
(e.g. as identified locally by the NPCT model) would likely
attract the 'near market' for cycling in these areas.
Areas with lower current levels of cycling may require greater
emphasis on complementary behavioural interventions in order to maximise the
value of infrastructure investement [@Sloman2014].

Across these categories, the current trip distance
distribution provides insight into the long-term potential rate of
cycling that could be expected. It is recommended that long-term
investment focusses on areas where many trips (or stages)
are of an easily cyclable distance.

## Short commutes and cycling

The proportion and absolute number of commutes that are less than
5 km in each LAD is illustrated in Fig. xx2.
Because the majority of new cycling trips are short, this data
can be used as a proxy for the potential level of cycling:
cycling is highly unlikely to become the dominant mode of transport
in areas where relatively few short-distance trips take place.
In areas where most trips are of a cyclable distance, by contrast,
cycling could in theory become "the first and natural choice"
for urban transport [@WelshGovernment2014].

In line with the 8 census distance bands used by the data distribution
portal *nomis* (with breaks at 2, 5 and 10 km) we defined 'short' trips
as those below 5 km in Euclidean distance. (Trips below 2 km were deemed
as 'very short'.) Fig. xx represents the
geographical distribution of these short trips for commuting in
absolute and proportional terms.

```{r, echo=FALSE, fig.cap="The geographical distribution of short (sub 5 km in Euclidean distance) trips in England."}
las$`Short trips (n)` <- las@data$All_.0.2. + las@data$All_.2.5.
las$`Short trips (%)` <- las$`Short trips (n)` / las@data$All_All * 100
qtm(las, fill = c("Short trips (n)", "Short trips (%)"), fill.n = 3, fill.palette = "Blues", fill.style = "jenks")
```

There is a correlation between the proportion of commutes in
an area that are short and the proportion of commutes that are
made by cycle (Fig. xx). Although the relationship is not
particularly strong ($R^2 = 0.17$), the plot is useful for
identifying the areas with a lower than expected rate of cycling
based on current trip distributions. Points below the line in
Fig. xx have disproportionately low level of cycling based on
commuting behaviour, indicating high future potential update.
The LADs with the largest negative residual have a much lower
rate of cycling than would be expected from the proportion of
short trips in the area. This is Hastings. The area with the
largest positive residual is Cambridge. These are illustrated
in Fig. xx with text.

The top 10 areas with the greatest unmet potential for cycling by
this measure are, in descending order,
Hastings, Blackburn with Darwen,
Burnley,                      Chesterfield,
Bradford,                     Oldham,
Hartlepool,                   Torbay,
Stoke-on-Trent  and             Calderdale.
Each of these local authorities are represented by a blue circle in
Fig. xx3.

```{r, echo=FALSE, fig.cap="Scatterplot illustrating the link between the proportion of commutes that are short and the proportion of commutes that are cycled."}
lmod1 <- lm(pcycle ~ `Short trips (%)`, data = las@data)
# summary(lmod1)
sel <- c(which.min(lmod1$residuals), which.max(lmod1$residuals))
sel2 <- head(order(lmod1$residuals), 10)
sel3 <- tail(order(lmod1$residuals), 10)
# las$NAME[sel3]
# library(cowplot)
ggplot(las@data, aes(`Short trips (%)`, pcycle)) + geom_point() +
  geom_abline(intercept = -0.922, slope = 0.1092, col = "red") +
  geom_text(data = las@data[sel,], aes(`Short trips (%)`, pcycle,  label = NAME)) +
  geom_point(data = las@data[sel,], aes(`Short trips (%)`, pcycle), color = "red") +
  geom_point(data = las@data[sel2,], aes(`Short trips (%)`, pcycle), shape = 1, size = 5, col = "blue") +
  ylab("Level of cycling (%)")
  # geom_line(aes(lmod1$fitted.values, `Short trips (%)`), col = "red")
las$short_trip_residual <- lmod1$residuals
# write.csv(las@data, "pct-bigdata/national/la-ttw-distance.csv")
# This clear pattern The relationship between the average *length* of trips and the current ...
```

The analysis suggests that cycling interventions in
such areas (with more than around 40% of trips being less than 5 km
yet a low rate of cycling) could, if effective, yield a high return on investment.
However, we note that the *cycling culture* of these areas is likely
to be weak and this may take time and 'soft' policy interventions,
in addition to high quality cycle infrastructure, to change.
Also, the example of South Cambridgeshire, in which only 21%
of commuting trips are 'short' yet the rate of cycling is high at
7.7%, demonstrates that
zones above the red line if Fig. x may also have
a high potential for cycling to
grow further. One potential
advantage of promoting cycling in an area that already
has a strong cycling community is that full use of
new infrastructure could be made more quickly than in other areas
[@Sloman2014].

The question of whether it is more effective to invest for cycling in
areas that are already 'cycling success stories' (as exemplified by the
Cycling Demonstration Towns project) or whether higher benefit-cost ratios
accrue in low cycling areas with a high potential (as exemplified by the
Leeds-Bradford 'CityConnect' investment) is an open question.
If anything, the analysis suggests that a wide range of area types
could benefit from cycling investment, especially those in which many
short distance which could switch to cycling take place.

# The national distribution of cycling potential

Building on the descriptive analysis presented above, this
section estimates cycling potential under the 'Cycling Delivery Plan'
scenario of cycling uptake, described in Appendix xx.
Applying the flow model to flows with distance distributions
representative of each authority allows for the scenarios described
in section xx to be applied nationwide. The results are presented
in Fig. 3 at the level of County and Unitary Authorities, the level
at which most transport investment is currently made.
Note that 'cycling potential' can be measured in
absolute or proportional terms and that authorities with a higher
population tend to have a higher absolute potential increase in cycling.
The distinction is important when allocating funding.

Cycling potential can be quantified as an absolute number of
trips/cyclists or as a proportion. Each metric has advantages.
The former is advantageous when allocating funding based on estimated
economic values of extra cycling activity. The latter is advantageous
when designing local targets or assessing whether local areas should
meet or exceed national-level targets. 

As with the number of short trips taking place in an area,
the absolute number of potential new cyclists is influenced by
the population of the local area: a large LAD will have a
greater cycling potential in absolute terms than a small LAD with the
same potential level (expressed as a %) of cycling.

Fig. x illustrates the geographical variability in the potential
increase in the rate of cycling under the Cycling Delivery Plan
Scenario. It is important to not that this is based *only* on the
trip distribution of commuter trips in each local authority;
hilliness is not taken into account. The values generated
correlate well with the observed level of cycling in the 2011
Census: $R^2 = 0.17$. Note that this is the same level of correlation
that was found between the percentage of short trips and the level
of cycling presented in the previous. 

```{r, echo=FALSE, fig.cap="Geographical variability in cycling potential accross England in percentage (left) and absolute (right) terms."}
las$pcycle_cdp <- las$pcycle + las$pcycle_exp
las$Bicycle_exp <- las$All_All * las$pcycle_exp / 1000
fixedBreaks2 <- c(0,2,3,5)
fixedBreaks3 <- c(0,2,3,5)
qtm(las, c("pcycle_exp", "Bicycle_exp"),
  fill.style = c("fixed", "quantile"),
  fill.n = 3,
  fill.breaks = fixedBreaks2, fill.title = c("Growth in cycling (%)",
    "Growth in cycling\n(1000 commuters)"))
```


The advantage of estimating the expected rate of cycling in this
way is that it estimates growth in cycling in absolute terms.
This could be useful for allocating funding between different Local
Authorities. The top 10 Local Authorities in terms of increased
number of commuter cyclists under our Cycling Delivery Plan is
presented in Table xx.

```{r, echo=FALSE}
t0.1 <- tail(las@data[order(las$Bicycle_exp),], 10)

t0.1 <- select(t0.1, Name = NAME, `Current %` = pcycle, `Growth %` = pcycle_exp, `Current (n)` = Bicycle_All, `Growth (n)` = Bicycle_exp, `% short car` = shortcar)
t0.1 <- t0.1[10:1,]
kable(t0.1, digits = 2, row.names = F)
```

Aggregated to the level of County and Unitary Authorities (CUAs), the level
at which many transport decisions are made, the top ten areas are
presented in Fig. x. The top 10 CUAs in terms of estimated potential
growth in cycling under our implementation of the Cycling Delivery Plan
are: Lancashire, Hampshire, Kent, Essex, Birmingham, Hertfordshire, Surrey,
Leeds, Staffordshire and Nottinghamshire.
 
Note that the size of these CUAs plays a strong role in
their potential for increased cyclists in absolute terms.
The 10 CUAs listed above have an average employed population
of more than 500,000. The average number of working people in all
CUAs, by contrast, is 170,000.
(For comparison, the average per LAD is just under 80,000). These numbers underline the importance
of measuring expenditure in pounds per person for areas, rather than
simply estimating the overall total.

```{r, echo=FALSE, fig.cap="Potential growth in cycling at the level of highway authorities in England", message=FALSE}
# Do for cuas too?
library(rgdal)
cuas <- raster::shapefile("pct-bigdata/national/cuas.shp")
lasp <- rgeos::gCentroid(las, byid = T)
ldf <- las@data
output_data <- select(ldf, Bicycle_exp)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$Bicycle_exp <- cuasp@data$Bicycle_exp
qtm(cuas, "Bicycle_exp", fill.title = "Growth in commuter cyclists (n)")
# cuas@data$CTYUA12NM[order(cuas@data$Bicycle_exp, decreasing = T)]
output_data <- select(ldf, All_All)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$All <- cuasp@data$All_All
# summary(cuas$All)
# summary(head(cuas@data$All[order(cuas@data$Bicycle_exp, decreasing = T)], 10))
```


# Method and discussion

The results presented in this paper are designed to provide a first
approximation of how the level of cycling that is attainable varies
across different
LADs in England depending on current trip distance distributions.
We are confident that the fundamentals of the
analysis are robust, based on very reliable
census data on the distances that
people travel to work at relatively high levels of geographical disaggregation.

The work underlines the fact that it is not feasible for a commuter
belt such as Sevenoaks, where only 19% of commuter trips are below
5 km, to have a very high rate of cycling to work.^[Integrating cycling with public transport in such areas, e.g. by
encouraging cycling to bus and train stations, may allow a higher
proportion of trip stages to be made by bicycle in such areas, however.
]
These are the fundamentals that should drive strategic
investments and policies promoting active travel.
That is not to say that cycling investment is
never justified in areas such as Sevenoaks: there may,
for example, be a higher
latent demand for leisure trips. However, the analysis
clearly suggests that areas with a high proportion of short trips,
based 2011 commuting behaviour,
which is relatively resistant to change, would benefit from
new investment in cycling. Identifying where such 'quick win'
areas are is an important step towards creating a nationally
'joined-up' cycling strategy and allocating funding based on
evidence.

There are a number of ways that the analysis could be extended,
including:

- Presenting the rate of cycling per 'short trip' instead of per trip
of any length - this would highlight areas that have done well despite
local trip-distance distributions.
- The addition of hilliness per LA as a predictor of the potential level of cycling.
- Using weather-based variables, such as number of rainy days per year, as explanatory variables.
- The use of transport infrastructure variables, such as the proportion
of population living within a certain distance of motorways and A roads, as explanatory variables.

The addition of additional
explanatory variables should improve the predictive
capability of the model. However, it is important to note that the
purpose of the model is not to predict the current rate of cycling,
but to estimate what the plausible rate of cycling could be under
scenarios of the future to guide strategic active travel policies.
An additional extension of this national-level
analysis would therefore be the estimation of cycling potential under
different future scenarios, other than the Cycling Delivery Plan. The
'Gender Equality', 'Go Dutch' and 'Ebikes' scenarios, used in the
flow-level model, would be logical extensions to the analysis in this
direction.

A wider point is that when the flow model described in Appendix xx
is deployed nationwide, model-based outputs per CUA would contain
estimates of the future level of cycling under each of the aforementioned
scenarios. This would minimise duplication in the workflow and integrate
the analysis of the flow data and the national data presented in this paper.

A question related to funding is the level at which cycling
investment will be allocated. If cycling policies are to form
part of an integrated national transport strategy, it is recommended
that funding for cycling is allocated at the level at which transport
investment for other modes takes place. Currently this is the level
of Combined and Unitary Authorities (CUAs) and, increasingly,
Combined Authorities and Local Enterprise Partnerships (LEPs).
CUAs with high unmet potential for cycling include
Lancashire, Hampshire and Kent. Our analysis suggests that
funding for cycling should move away from
sporadic and time-limited city-by-city investment to
a national-level strategy. Long-term funding for every Highway
Authority dedicated to cycling will enable a joined-up
and cost-effective cycling network to be created nationwide.

# References

```{r, echo=FALSE, eval=FALSE}
file.copy("documents/national.docx", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.docx", overwrite = T)
file.copy("documents/national.pdf", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.pdf", overwrite = T)
```

