---
title: "National-level analysis of cycling potential"
output: pdf_document
---

```{r, echo=FALSE}
pkgs <- c("grid", "png", "knitr", "tmap")
lapply(pkgs, library, character.only = T)
opts_knit$set(root.dir = "../")
# source("analysis/la-analysis.R", print.eval = T)
```

```{r}
las <- readRDS("pct-bigdata/national/las.Rds")
```



# Introduction



Cycling potential operates at many levels. At the lowest level,
the decision of whether or not to cycle depends on personal
and highly localised factors. Is my cycle in good working order?
Is my body in good working order? Is the weather nice outside?

On the other hand, larger-scale process are at play to determine the
proportion of trips that are cycled. The terrain, roads, traffic and
local culture will all conspire to make more or fewer trips cycled
in an area compared with national and international averages.

Critical hear is the distribution of trip lengths associated with
a particular area. In some 'commuter belts' almost all trips are
long-distance, precluding a strong utility cycling culture.
In vibrant urban centres, by contrast, a large proportion of
travel is within cycling distance. This helps explain why the rate
of cycling to work (and likely cycling overall too)
has grown in large cities such as London and Manchester, but remained
largely stagnant at the national level.

In this paper we explore the geographical variability of
*cycling potential* at the national level.
To do so, we must first describe briefly the current rate of cycling
nationwide. The region of analysis is the entirety of England and,
although now slightly out of date, it focussed on travel
to work. Commuting data from the 2011 Census
has the greatest coverage (90% + of the working population) and
geographical resolution and coverage (with a high response rate from
every authority) of any travel dataset of England. Commuting data
also reflect utility cycling, as opposed to leisure cycling, which is
the focus of the National Propensity to Cycle Tool project, of which
this analysis is a part.

The unit of analysis is the Local

The structure of the paper is as follows. First we will describe the
current rate of cycling. The subsequent section will explore the potential
for additional cycle trips at the national level, both in terms of
the proportion of all commuter trips cycled and in terms of the
absolute number of commuter cyclists. (The former can provide an
evidence base, amongst others, to inform the setting of suitable local
cycling targets; the latter can assist with the relative allocation of
funding for cycling investment between different local authorities.)
Finally we explain the methodology that was used and the impact that this
has on the policy implications of the work.

# The current level of cycling

The current rate of cycling in England is depicted in Fig. 1.
This shows that the spatial distribution of commuter cycling is
highly uneven. In 135 LADs (42% of the total number and population)
less than 2% of workers report cycling as their main means of travel
to work. Nearly half (160) of the 324 LADs have an intermediate level
of cycling, between 2 and 5%. 
The remaining 29 LADs have what can be considered a high rate of
cycling by English standards of over 5% of commuters using a cycle.
This means that only around 10% of the population can be considered
as living in areas where cycling is a 'normal' mode of travel to work.

```{r, echo=FALSE, fig.cap="The current level of cycling to work in England from the 2011 Census."}
fixedBreaks = c(0,2,5,30)
qtm(las, "pcycle", fill.style = "fixed", fill.breaks = fixedBreaks)
```

The top 10 LADs by cycling share in commuting is presented in Table 1.


|    |Name                        | Current %| Expected %| Current (n)| Expected (n)| % short car|
|:---|:---------------------------|---------:|----------:|-----------:|------------:|-----------:|
|58  |Cambridge                   |      28.9|        3.7|       17313|         2232|        16.1|
|36  |Oxford                      |      17.0|        3.4|       11962|         2395|        18.0|
|155 |Hackney                     |      13.8|        3.2|       16411|         3809|        10.7|
|249 |York                        |      11.2|        3.6|       11143|         3522|        24.3|
|273 |Gosport                     |      10.7|        3.3|        4296|         1347|        40.7|
|186 |Islington                   |       9.1|        3.7|        9775|         3960|         6.4|
|5   |Norwich                     |       8.7|        4.0|        5480|         2523|        24.0|
|226 |Kingston upon Hull, City of |       8.1|        4.0|        8974|         4405|        35.6|
|306 |Lambeth                     |       7.7|        3.0|       12945|         4966|        13.0|
|140 |South Cambridgeshire        |       7.6|        1.9|        6041|         1537|        35.4|