---
title: "National-level analysis of cycling potential"
output: word_document
---

```{r, include=FALSE}
pkgs <- c("grid", "png", "knitr", "tmap", "dplyr")
lapply(pkgs, library, character.only = T)
opts_knit$set(root.dir = "../")
# source("analysis/la-analysis.R", print.eval = T)
```

```{r, echo=FALSE}
las <- readRDS("pct-bigdata/national/las.Rds")
# source("analysis/las-tables.R")
```



# Introduction

Cycling potential operates at many levels. At the lowest level,
the decision of whether or not to cycle depends on personal
and highly localised factors. Is my cycle in good working order?
Is my body in good working order? Is the weather nice outside?
On the other hand, larger-scale phenomena are at play to determine the
proportion of trips that are cycled. The terrain, roads, traffic and
local culture will all conspire to make more or fewer trips cycled
in an area compared with national and international averages.

In this context, the distribution of trip lengths associated with
a particular area is critical to its cycling potential.
In some 'commuter belts' almost all trips are
long-distance, precluding a strong utility cycling culture.
In vibrant urban centres, by contrast, a large proportion of
travel is within cycling distance. This helps explain why the rate
of cycling to work (and likely cycling overall too)
has grown in large cities such as London and Manchester, but remained
largely stagnant at the national level.

In this paper we explore the geographical variability of
*cycling potential* at the national level.
To do so, we must first describe briefly the current rate of cycling
nationwide. The region of analysis is the entirety of England and,
although now slightly out of date, it focussed on travel
to work. Commuting data from the 2011 Census
has the greatest coverage (90% + of the working population) and
geographical resolution and coverage (with a high response rate from
every authority) of any travel dataset of England. Commuting data
also reflect utility cycling, as opposed to leisure cycling, which is
the focus of the National Propensity to Cycle Tool project, of which
this analysis is a part.
The unit of analysis is the 324 Local Authority Districts in England,
the highest resolution data that provides a breakdown of mode by distance.

The structure of the paper is as follows. First we will describe the
current rate of cycling. The subsequent section will explore the potential
for additional cycle trips at the national level, both in terms of
the proportion of all commuter trips cycled and in terms of the
absolute number of commuter cyclists. (The former can provide an
evidence base, amongst others, to inform the setting of suitable local
cycling targets; the latter can assist with the relative allocation of
funding for cycling investment between different local authorities.)
Finally we explain the methodology that was used and the impact that this
has on the policy implications of the work.

# The current level of cycling

The spatial distribution of of commuter cycling across England
is depicted in Fig. 1. The dataset shows that there is wide variation
in the level of cycling, which can be split roughly into
3 categories. In 135 LADs (42% of the total number and population)
less than 2% of workers report cycling as their main means of travel
to work. Nearly half (160) of the 324 LADs have an intermediate level
of cycling, between 2 and 5%. 
The remaining 29 LADs have what can be considered a high rate of
cycling by English standards, of 5% of more. 
This means that only around 10% of the population can be considered
as living in areas where cycling is a 'normal' mode of travel to work.
8 of the 'high cycling' LADs are
based in London.^[These are: Hammersmith and Fulham,
Wandsworth,             Hackney, Tower Hamlets,
Islington,              Richmond upon Thames,
Southwark and              Lambeth.]


```{r, echo=FALSE, fig.cap="The current level of cycling to work in England from the 2011 Census."}
fixedBreaks = c(0,2,5,30)
qtm(las, "pcycle", fill.style = "fixed",
  fill.breaks = fixedBreaks, fill.title = "Percent cycling")
```

Outside London, 'high cycling' LADs tend to be high density urban
areas focussed on a large mono-centric centre. Cambridge (with 29% cycling),
Oxford (17%) and York (11%) are well known high-cycling Local Authorities
that fit this category but the list also contains lesser-known
areas. Norwich (9%), Bristol (7%), Lincoln (6%) and Portsmouth and
Gosport (with 7 and 10% cycling to work, respectively) also have
a relatively high rate of cycling. 
Further details on the top 10 LADs by cycling share in commuting
are provided Table 1.

Table: The top 10 'high cycling' Local Authorities, in terms of the
proportion of commutes made by bicycle, from the 2011 Census.

```{r, echo=FALSE}
t0 <- tail(las@data[order(las$pcycle),], 10)
t0 <- t0[10:1,]

t0 <- select(t0, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)
kable(t0, digits = 1, row.names = F)
```

It is equally interesting to consider areas with the lowest
rates of cycling; these are presented in Table 2. It is interesting
to note that these are not exclusively rural areas, where low cycling could
be explained by the lack of jobs near to houses.
Five of the 10 Local Authorities with the lowest level of cycling to work
are largely rural: Rossendale, Harrow, Tandridge,
Bolsover and North East Derbyshire.
These are by no means the *most* rural areas in
England and, furthermore, it is interesting to note that each
lies in close proximity to Motorways and other
large roads.
Bradford seems like an outlier in this group as it is
known primarily for its well-defined city centre. However,
the Local Authority also contains large expanses of rural areas,
so can to some extent be grouped with the aforementioned areas.

Two of the areas in the 'bottom 10' for commuter cycling are
commuter belts: Sevenoaks and Epping Forest are wealthy areas which lie
on the outskirts of London. 
The remaining three areas are
relatively deprived urban areas that are part of a
wider conurbation. Some of these areas
(Barnesley, ) have centres that are
car-dominated.

Table: The bottom 10 'high cycling' Local Authorities, in terms of the
proportion of commutes made by bicycle, from the 2011 Census.

```{r, echo=FALSE}
t0.1 <- head(las@data[order(las$pcycle),], 10)

t0.1 <- select(t0.1, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)

kable(t0.1, digits = 2, row.names = F)
```

This clear pattern
The relationship between the average *length* of trips and the current ...

# The national distribution of cycling potential

Cycling potential can be quantified as an absolute number of
trips/cyclists or as a proportion. Each metric has advantages.
The former is advantageous when allocating funding based on estimated
economic values of extra cycling activity. The latter is advantageous
when designing local targets or assessing whether local areas should
meet or exceed national-level targets. 

Of course, the absolute number of potential new cyclists is influenced by
the population of the local area: a large LAD will have a
greater cycling potential in absolute terms than a small LAD with the
same potential level (expressed as a %) of cycling.

Because the majority of new cycling trips are short, the
number/proportion of trips in an area that are short can be used as
a proxy for the potential level of cycling.
In line with the 8census distance bands used by the data distribution
portal *nomis* (with breaks at 2, 5 and 10 km) we defined 'short' trips
as those below 2 km in Euclidean distance. (Trips below 2 km were deemed
as 'very short'.)
Fig. xx represents the
geographical distribution of these short trips for commuting in
absolute and proportional terms.

```{r, echo=FALSE, fig.cap="The geographical distribution of short (sub 5 km in Euclidean distance) trips in England."}
las$`Short trips (n)` <- las@data$All_.0.2. + las@data$All_.2.5.
las$`Short trips (%)` <- las$`Short trips (n)` / las@data$All_All * 100
qtm(las, fill = c("Short trips (n)", "Short trips (%)"), fill.n = 3, fill.palette = "Blues", fill.style = "jenks")
```

There is a correlation between the proportion of commutes in
an area that are short and the proportion of commutes that are
made by cycle (Fig. xx). Although the relationship is not
particularly strong ($R^2 = 0.17$), the plot is useful for
identifying the areas with a lower than expected rate of cycling
based on current trip distributions. Points below the line in
Fig. xx have disproportionately low level of cycling based on
commuting behaviour, indicating high future potential update.

```{r, echo=FALSE, fig.cap="Scatterplot illustrating the link between the proportion of commutes that are short and the proportion of commutes that are cycled."}
lmod1 <- lm(pcycle ~ `Short trips (%)`, data = las@data)
# summary(lmod1)
library(cowplot)
ggplot(las@data, aes(`Short trips (%)`, pcycle)) + geom_point() +
  geom_abline(intercept = -0.922, slope = 0.1092, col = "red") +
  ylab("Level of cycling (%)")
  # geom_line(aes(lmod1$fitted.values, `Short trips (%)`), col = "red")
las$short_trip_residual <- lmod1$residuals
# write.csv(las@data, "pct-bigdata/national/la-ttw-distance.csv")
```





# Method and discussion

The results presented in this paper are designed to provide a first
approximation of the level of cycling that is attainable in different
LADs across England. We care confident that the fundamentals of the
analysis are robust, based on census data on the distances that
people travel to work. It is clearly not feasible for a commuter
belt such as Sevenoaks, where only 19% of commuter trips are below
5 km, to ever have a high level of cycling. This is captured in the
analysis. That is not to say that cycling investment is
never justified in areas such as Sevenoaks: there may,
for example, be a higher
latent demand for leisure trips. However, the analysis
clearly suggests that areas with a high proportion of short trips,
based 2011 commuting behaviour,
which is relatively resistent to change, would benefit from
new investment in cycling. Identifying where such 'quick win'
areas are is an important step towards creating a nationally
'joined-up' cycling strategy and allocating funding based on
evidence.

There are a number of ways that the analysis could be extended,
including:

```{r, echo=FALSE}
file.copy("documents/national.docx", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.docx", overwrite = T)
file.copy("documents/national.pdf", "~/Dropbox/DfT bid/Draft Writeups/Robin/national.pdf", overwrite = T)
```

