---
title: 'Estimating propensity to cycle in Sao Paulo: a flow-level model and interactive
  tool'
author: "Robin Lovelace"
date: "April 12, 2015"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  word_document: default
bibliography: ~/Documents/Transport.bib
---

```{r, echo=FALSE, include=FALSE}
pkgs <- c("stplanr", "knitr", "xtable", "stplanr", "sp", "grid", "png")
vapply(pkgs, require, character.only = T, FUN.VALUE = logical(1))
```


# Introduction

This proposal sets out an agenda to implement, refine and further
develop work begun in the UK to create a National Propensity to Cycle Tool
(NPCT). Funded by the English Department for Transport
(DfT), the NPCT project aims to provide new evidence on the question of 
*where* new sustainable transport interventions should be prioritised.
The flagship output of the NPCT is an interactive on-line planning
tool, which provides real-time feedback to transport planners for exploring
current rates of cycling and where this could be boosted at the city level
(Fig. 1). During user testing at the DfT's Offices on Friday 24^th^
April, professional transport planners had the opportunity to test
a live version of the tool and provide feedback. Overall the feedback was
very encouraging, indicating that there is high demand for the tool in
Local Authorities across the country. 

```{r, echo=FALSE, fig.cap="Prototype of the NPCT used for user testing at the Department for Transport in April. Note the flows represent model-based estimates of where there is greatest potential for increased cycling, based on official data."}
grid.raster(readPNG("~/Dropbox/DfT bid/figures/user-tested-version.png"))
```

This document sets out a research plan to deploy the NPCT methodology
in Soa Paulo. The work can be justified on academic, methodological and
policy grounds.

Specifically, the research seeks to identify travel corridors and routes
along the transport network where new **bicycle paths** would
have a greatest positive impact on citizens. The focus
on infrastructure at the city level is timely because of the rapid
roll-out of pro-cycling interventions worldwide, primarily in terms of bicycle
paths and bicycle share schemes
[@OBrien2013]. In Latin America alone,
substantial capital injections over the last 5 years have boosted cycling in
many cities, including
Bogotá, Rio de Janeiro and Sao Paulo [@Mosquera2012; @Jones2013b].
However, the speed of investment
meant that there has been limited time to decide precisely where the new
network should be place to maximise benefit-cost ratios. 

From an *academic perspective*, the project will fill a research gap in the transport
evaluation and modelling literature.
Active travel research has traditionally focussed on overall cost:benefit ratios.
The impact of the location of new infrastructure on these ratios has
been relatively under-studied, especially in the context of active travel.

*Methodologically*, the project will also break new ground as its main output
will be an interactive on-line map, available to transport planners
(and eventually the public) to assist with open decision-making processes.
Specifically, the use of the recently developed R package **shiny** in tandem
with the JavaScript library **leaflet** is a first in the field.

The extension of the NPCT project to Sao Paulo has substantial
*policy implications*. In the long-run, demonstration of the methodology in
a developing world context, in addition to its utilisation in the West, should
help with the deployment of interactive on-line transport planning tools
for improved active travel provision worldwide. There are strong reasons for
choosing Sao Paulo as the next place outside of Europe for the tool's deployment.
There is political momentum behind the idea of
increased active travel in Sao Paulo and evidence that
walking and cycling is associated with health benefits from a sample of
3 cities in Brazil [@Reis2013].

However, there has also been
opposition to the construction of bicycle paths.
Part of this backlash has been the perception
that the process of planning the new paths has been rushed.^[See
[folha.uol.com.br/internacional/en/saopaulo/](http://www1.folha.uol.com.br/internacional/en/saopaulo/2015/03/1605612-sao-paulo-court-of-justice-orders-mayor-haddad-to-paralyse-bike-paths-under-construction.shtml), for example.]

There is clearly a need for local authorities to strategically plan and
justify future pro-walking and cycling investments. The NPCT method is
ideally suited to meet this need.

# Data

The basic model requires only two sets of input data
(although more refined versions benefit from many additional datasets):

- *Flow data* estimating the rate of movement between different places,
often recorded through Census travel to work questions. The data is
generally provided as a 'flow matrix' (with rows representing origins
and columns destinations) or a longer table of origin-destination pairs.
This type of data is increasingly available from so-called 'Big Data'
sources such as passively collected mobile telephone records.

- *Geographical data* on the coordinates of trip origins and destinations.
As a bare minimum, this means the (preferably population-weighted)
centroids of each zone in the study area but could also include.

The flow model described in this paper can work for
anywhere that has access to these two data types. To link
the two datasets together, *zone ids* are needed in both
datasets. To ease the process of combining *flow* and *point* data,
an R function called `gFlow2Lines()` was 
created.^[The source code of `gFlow2Lines()` has been made available
on-line, as part of an R package for sustainable transport planning,
**stplanr**. See [github.com/Robinlovelace/stplanr/](https://github.com/Robinlovelace/stplanr/blob/master/R/gFlow.R).]
The two input datasets and single output, a set of geographically defined
lines with attributes for each flow in both directions
(which we label `flowlines`) are illustrated in the tables (1 and 2)
and Fig. 2 below.

```{r tflow, echo=FALSE, results='asis'}
t1 <- xtable(head(flow[c(1:3, 12)]), caption = "Sample of the 'flow' input dataset, representing the number of people who commute from locations within and between administrative zones (MSOAs)", label = "tbl:flow")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r tcents, echo=FALSE, results='asis'}
t2 <- xtable(as.data.frame(cents[1:3,-c(3,4)]), caption = "Sample of the 'cents' input dataset, representing the geographical location of the population-weighted centroids of MSOA zones described in Table~\\ref{tbl:flow}", label = "tbl:cents")
print(t1, type = "latex", comment = FALSE, caption.placement = "top")
```

```{r, echo=FALSE, fig.cap="Illustration of 'flow data' converted into geographical lines between origin and destination pairs. Width represents the total number of trips. Note the circles, which represent intra-zone flow.", fig.height=4}
data("flowlines")
plot(flowlines, lwd = flowlines$All / 10)
```


Although the details may differ (e.g. the categories used to disaggregate flow
by vehicle mode, trip type and socio-demographic group), the basic structure
of flow data will be the same the world over.

## Count data

An issue with the NPCT approach at present is lack of verification. We cannot
be sure that the origin-destination combinations that the model identifies
as having the highest potential rate of cycling really will respond best to
pro-cycling interventions. Also, there is no way of know whether our attempts
to allocate 'fastest' and 'quietest' routes to the travel network are
accurate. There is a great need for testing estimates of *route-allocated*
propensity against real-world data.


There is therefore a need for cycle path count data at,
specific points on the network in the model, to improve test and subsequently
improve the model's ability to allocate cycling potential to the transport
network. 

# Method

Underlying the estimates of increased cycling level is a General Linear regression
Model (GLM), which uses a number of explanatory variables to estimate the expected
proportion of trips that *should* be made between all OD pairs in the study area.
At present distance, hilliness, age distribution and gender split of trips are
the input variables. This model could be expanded
to include the quality of
transport network, the level of car-use and the accident rate.

The flows are allocated to the travel network using the Application Programming
Interface (API) of
[CycleStreets.net](http://www.cyclestreets.net/api/), an on-line journey planner
developed by cyclists for cyclists. A unique feature of this method is that it
allows routes to be allocated for the *fastest*, *quietest* and *most balanced*
journeys. This is useful for transport planners seeking to assess on which
roads new cycle paths are most urgently needed (using the fastest route)
and how much of a detour cyclists must currently take to find a 'quiet' path
(assuming there is a quiet path available at all). 

Although the CycleStreets.net API currently works only for the UK and parts of
continental Europe, it is open software and based on OpenStreetMap data. A version
could be created for Brazil. Alternatively, a third-part route planner such as
OSRM^[See http://project-osrm.org/] or GraphHopper^[See https://graphhopper.com/]
could be used.

Most of the code underlying the NPCT project is open source, so the method is
easily reproducible.^[See https://github.com/npct] In addition, an R package is
under development to help automate some of the data processing and
analysis that underlies the NPCT methodology.^[See https://github.com/Robinlovelace/stplanr]

# Proposal

Based on the above, it is proposed that Dr Robin Lovelace visits the
MobiLab research centre in Sao Paulo for a 1 month placement. This will allow
for the development of an NPCT method for Sao Paulo and potentially other
cities in Latin America. The key outputs of this placement will be:

- Development and deployment of an NPCT for Sao Paulo, ready for use by transport
planners working to improve active travel in the city.
- Development of software and methods that will enable the NPCT approach to
be deployed in other cities across Brazil, Latin America and potentially worldwide.
- Write-up of the academic research, including papers on the methodology,
software and findings comparing how propensity to cycle varies geographically
from one city to another.

# References
