---
title: "Travel to work areas"
author: "Robin Lovelace"
date: "February 21, 2015"
output: pdf_document
---
```{r setup, include=FALSE}
library("knitr")
opts_knit$set(root.dir = "../")
```

```{r, echo=FALSE, include=FALSE}
source("set-up.R")
pkgs <- c("raster")
lapply(pkgs, library, character.only = TRUE)
```

Travel to work areas (TTWA) are geographic zones in which
"where the bulk of the resident population also work within the same area".
They can be seen as 'commuter watersheds', and are hence useful for analysing
commuter patterns. Because estimates of distance decay depend strongly on
an understanding of commuter zones, TTWA may be useful 'train' DD paramaters
for relatively similar commuting blocks.

TTWA boundaries can be downloaded by the public directly from the internet:

```{r, echo=FALSE}
# url = "https://geoportal.statistics.gov.uk/Docs/Boundaries/Travel_to_work_areas_%28E+W%29_2007_Boundaries_%28Generalised_Clipped%29.zip"
# download.file(url, method = "curl", destfile = "bigdata/ttwa.zip")
# unzip(zipfile = "bigdata/ttwa.zip", exdir = "bigdata/")
```

To load the files one can use the `shapefile` function of **raster**.
The `gMapshape` function intelligently simplifies the boundaries.
There are 186 TTWA in England and Wales.

```{r}
# gMapshape("bigdata/TTWA_DEC_2007_EW_BGC.shp", percent = 5)
ttw <- shapefile("bigdata/TTWA_DEC_2007_EW_BGCmapshaped_5%.shp")

head(ttw@data, 3)
nrow(ttw)
# plot(ttw)
```

# Areas of interest

Let's highlight the areas we are interested in:

```{r}
aoi <- "Manchester|Coventry|Norwich"
sel <-  grep(aoi, ttw$TTWA07NM)
plot(ttw)
plot(ttw[sel,], col = "red", add = T)
```

Let's take a look at each of these in order.
In the plots below, the grey areas with a thick black boundary are TTWA.
Thin black lines with no fill represent MSOA areas in the wider area
(where present). The red zones represent the city Local Authorities,
such as Norwich.

```{r, echo=FALSE}
source("geo-load/load-uk.R")
```


## Manchester

The size of the Manchester TTWA is roughly the same as Greater Manchester
Authority, but a different shape:

```{r}
plot(ttw[sel[2], ], lwd = 5, col = "grey")
plot(gman, add = T)
plot(man, add = T, col = "red")
```

## Coventry

Coventry's TTWA is substantially larger than the city bounds, extending
far out to the north:

```{r}
plot(ttw[sel[1], ], lwd = 5, col = "grey")
plot(cov, add = T, col = "red")
```

## Norwich

Norwich's TTWA is roughly the same size and shap as Norwich Local Authority
plus the Broadland and South Norfolk, the local authorities to the north
and south of it respectively.

```{r}
plot(ttw[sel[3], ], lwd = 5, col = "grey")
plot(wnor, add = T)
plot(nor, add = T, col = "red")
```




