---
title: "NPCT model results"
author: "Robin Lovelace"
output: 
  html_document: 
    fig_caption: yes
---

```{r, include=FALSE}
start_time <- Sys.time() # for timing the script
source("set-up.R") # pull in packages needed
library(scales)
```

This document was produced automatically at `r start_time`.

## Introduction

The results of NPCT scenarios are based on a model.
This document presents information about the input data, model diagnostics,
run time and information about its outputs.

The aim is to provide further information for transport planners and
researchers on the origin of results presented in the NPCT's interactive map.
This document is designed for use by advanced users: some technical knowledge
is needed to understand all its outputs.

The model driving the NPCT is licensed under the open source MIT License
and can be modified by others provided attribution to the original.

## Initial parameters

The preset values used to select the study area and filter the flow data
were as follows:

```{r, warning=FALSE}
# Set local authority and ttwa zone names
la <- "manchester" # name of the local authority
dir.create(paste0("pct-data/", la)) # on a unix machine

# Minimum flow between od pairs, subsetting lines. High means fewer lines.
mflow <- 30
mdist <- 10 # maximum euclidean distance (km) for subsetting lines
min_zones <- 60 # minumum number of zones in study area before buffer used
buff_dist <- 3 # buffer (km) used to select additional zones (often redundant)
```

## Input zone data

The input zones area are summarised in this section.

```{r plotzones, message=FALSE, warning=FALSE, results='hide'}
library(rgdal)
ukmsoas <- shapefile("pct-bigdata/national/msoas.shp")

# Load population-weighted centroids
cents <- readOGR("pct-bigdata/national/cents.geojson", layer = "OGRGeoJSON")

# Load local authorities and districts
las <- readOGR(dsn = "pct-bigdata/national/las-pcycle.geojson", layer = "OGRGeoJSON")
```


```{r, include=FALSE}
# lasdat <- SpatialPointsDataFrame(coords = coordinates(las), data = las@data)

# Load counties and unitary authorities
# cuas <- readOGR(dsn = "pct-bigdata/national/cuas.geojson", layer = "OGRGeoJSON")
# proj4string(lasdat) <- proj4string(las)
# cuas <- aggregate(lasdat, cuas, mean, na.action = na.omit()) # todo: fix data
# cua_shape <- cuas[grep(pattern = la, x = cuas@data$NAME)] # todo: fix
# # tmap::qtm(cuas2,fill = "clc")

la_shape <- las[grep(pattern = la, x = las@data$NAME, ignore.case = T),]
proj4string(la_shape) <- proj4string(ukmsoas)
cents <- spTransform(cents, CRSobj = proj4string(ukmsoas))
cents_la <- cents[la_shape,]
zones <- ukmsoas[cents_la, ]
zbuf <- gBuffer(zones, width = buff_dist * 1000) # create la zone outline

# Extract zones to plot
# zones <- ukmsoas[ grep(la, ukmsoas$geo_label), ] # extract by name

if(nrow(zones) < min_zones){
  cents_la <- cents[zbuf, ]
  zones <- ukmsoas[cents_la, ]
}

cents <- cents_la
```

The characteristics of zones are as follows:

```{r, echo=FALSE}
nzones <- nrow(zones) # how many zones?
mzarea <- round(median(gArea(zones, byid = T) / 10000), 2) # average area of zones, sq km
```

- Number of zones: `r nzones`, compared with 6791 in England
- Median area of zones: `r mzarea` ha, compared with 300 ha in the UK

## Input flow data

```{r, echo=FALSE}
flow <- readRDS("pct-bigdata/national/flow_eng_avlslope.Rds")

# Subset by zones in the study area
o <- flow$Area.of.residence %in% cents$geo_code
d <- flow$Area.of.workplace %in% cents$geo_code
flow <- flow[o & d, ] # subset flows with o and d in study area
sel <- flow$All > mflow # subset flows by n. people using it
```

```{r, echo=FALSE}
# nrow(flow) # how many OD pairs in the study area?
# proportion of flows in min-flow based subset
pmflow <- round(sum(sel) / nrow(flow) * 100, 2)
# % all trips covered
pmflowa <- round(sum(flow$All[sel]) / sum(flow$All) * 100, 2)
```

There are **`r nrow(flow)`** flows with origins and destinations in the study
area. Of these, **`r sum(sel)`** meet the criteria that `r mflow` people travel
on flow on a regular basis (a modifiable parameter used
to reduce the computational requirements
of the model). That's  **`r pmflow`%** for flow lines, accounting for
**`r pmflowa`%** of commuters in the study area.

```{r distance-dist, echo=FALSE, fig.cap="The study area (dark line), selected zones for training the model (grey) and the administrative zone of interest (red line). The black straight lines represent the most intensive commuting flows.", echo=FALSE, message=FALSE, warning=FALSE}
flow <- flow[sel, ]
# nrow(flow) # new flow rate

flow$id <- paste(flow$Area.of.residence, flow$Area.of.workplace)

l <- gFlow2line(flow = flow, zones = cents)
plot(zones, col = "lightgrey")
plot(las, add = T)
plot(zbuf, lwd = 5, add = T)
plot(la_shape, border = "red", add = T, lwd = 3)
lines(l[l$All > 100,])
```

```{r, echo=FALSE}
proj4string(l) <- proj4string(cents)
l$dist <- gLength(l, byid = T) / 1000 # Euclidean distance
dsel <- l$dist < mdist

l <- l[dsel,]
l <- l[l$dist > 0, ] # to remove flows of 0 length
```

## Hilliness of flows

The average hilliness of zones in the study area is
`r round(mean(zones$avslope), 2)`
degrees,
compared with the national average of 
`r round(mean(ukmsoas$avslope, na.rm = T), 2)`. This data is displayed in the
figure below.

```{r, echo = FALSE}
tm_shape(zones) +
  tm_fill("avslope", n = 3, palette = "Oranges")
```

```{r, echo=FALSE}
# Hilliness of flows
# (calculated as the average gradient of the zone
# of the flow's origin and destination, in degrees)
# is 
# `r round(mean(flow$avslope * flow$All / mean(flow$All), na.rm = T), 2)`.
# The UK
# average is xx degrees
```

## Lines allocated to the road network

We use CycleStreets.net to
estimate optimal routes. 
An illustration of these routes is presented below.

```{r flow-vars, echo=FALSE}
# # # # # # # # # # # # # # # # # #
# Calculate flow-level variables: #
# distances and olc for ag. model #
# # # # # # # # # # # # # # # # # #

# Calculate distances (eventually use route distance)


# Transform CRS to WGS84 for plotting
zones <- spTransform(zones, CRS("+init=epsg:4326"))
cents <- spTransform(cents, CRS("+init=epsg:4326"))
l <- spTransform(l, CRS("+init=epsg:4326"))

# # # # # # # # # # # # # # #
# Allocate flows to network #
# Warning: time-consuming!  #
# Needs CycleStreet.net API #
# # # # # # # # # # # # # # #

# Create local version of lines; if there are too many in the TTWA, sample!
l_local_sel <- l@data$Area.of.residence %in% zones$geo_code &
  l@data$Area.of.workplace %in% zones$geo_code
if(nrow(l) > 2 * sum(l_local_sel) & nrow(l) > 5000){ # sample if too many lines
  l_all <- l
  f <- list.files(paste0("pct-data/", la, "/"))
  if(sum(grepl("l_all", f)) == 0) saveRDS(l, paste0("pct-data/", la, "/l_all.Rds"))
#   l <- readRDS(paste0("pct-data/", la, "/l_all.Rds")) # restart point
  set.seed(2050)
  # sample from all routes in the TTWZ - change 1 for different % outside zone
  lsel <- sample(which(!l_local_sel), size = sum(l_local_sel) * 1)
  lsel <- c(lsel, which(l_local_sel))
  length(lsel)
  l <- l_all[lsel, ] # subset the lines
  # plot(l)
  # lines(l[2000:2600,], col = "blue") # ensure we have all the local ones
}

# Create route allocated lines
if(length(grep("rf.Rds|rq.Rds", list.files(paste0("pct-data/", la)))) >= 2){
  rf <- readRDS(paste0("pct-data/", la, "/rf.Rds")) # if you've loaded them
  rq <- readRDS(paste0("pct-data/", la, "/rq.Rds"))
  l <- readRDS(paste0("pct-data/", la, "/l.Rds"))
} else{
  rf <- gLines2CyclePath(l[ l$dist > 0, ])
  rq <- gLines2CyclePath(l[ l$dist > 0, ], plan = "quietest")

  # Process route data
  rf$length <- rf$length / 1000
  rq$length <- rq$length / 1000
  saveRDS(rf, paste0("pct-data/", la, "/rf_ttwa.Rds")) # save the routes
  saveRDS(rq, paste0("pct-data/", la, "/rq_ttwa.Rds"))
  saveRDS(l, paste0("pct-data/", la, "/l.Rds"))
  }
```

```{r plot-rlines, echo = FALSE, fig.cap="Sample of the straight and route-lines allocated to the travel network"}
plot(l[1:300,])
lines(rf[1:300,], col = "red")
lines(rq[1:300,], col = "green")
```

```{r, echo=FALSE}

# Allocate route factors to flows
# nz <- which(l$dist > 0) # non-zero lengths = nz
l$dist_quiet <- l$dist_fast <- l$cirquity <- l$distq_f <- NA
l$dist_fast <- rf$length
l$dist_quiet <- rq$length
l$cirquity <- rf$length / l$dist
l$distq_f <- rq$length / rf$length
```

## Distance distributions

The distance distribution of trips in the study area is displayed in the figure below, which compares the result with the distribution of trips nationwide.

```{r, echo=FALSE, fig.cap="Distance distribution of all trips in study lines (blue) compared with national average (dotted bars)"}
luk <- readRDS("pct-bigdata/national/l_sam8.Rds")

hdfl <- dplyr::select(l@data, All, dist_fast)
hdfl$Scope <- "Local"
hdfl$All <- hdfl$All / sum(hdfl$All)

hdfu <- dplyr::select(luk@data, All, dist_fast)
hdfu$Scope <- "National"
hdfu$All <- hdfu$All / sum(hdfu$All)

histdf <- rbind(hdfl, hdfu)

ggplot(histdf) +
  geom_histogram(aes(dist_fast, weight = All, fill = Scope, linetype = Scope),
    position = "identity", colour = "black", binwidth = 0.5) +
  scale_fill_manual(values = c("blue", NA)) +
  scale_linetype(c(1, 2), guide = "none") +
  scale_y_continuous(labels = percent) + 
  xlab("Route distance (km)") + 
  ylab("Proportion of trips in each band (under 10 km)") + 
  xlim(c(0,13))

pl5kmuk <- round(sum(luk$All[luk$dist_fast < 5]) /
    sum(luk$All) * 100, 2)
pl5km <- round(sum(l$All[l$dist_fast < 5]) /
    sum(l$All) * 100, 2)
```

From the nationwide sample of trips, `r pl5kmuk`% of trips are less than 5km.

In the case study area
`r pl5km`% of sampled trips are less than 5km.

Subsetting by distance (set
to `r mdist` km) and removing inter-zone flows
further reduces the number of flows from `r sum(sel)`
to `r nrow(l)`.

```{r, echo=FALSE}
# a = 11
# plot(l[a,])
# lines(rf[a,], col = "red")
# lines(rq[a,], col = "green")

# # # # # # # # # # # # # #
# Estimates slc from olc  #
# # # # # # # # # # # # # #

l$clc <- l$Bicycle / l$All
flow_ttwa <- flow # save flows for the ttwa
flow <- l@data
```

## The flow model

To estimate the potential rate of cycling under different scenarios
regression models operating at the flow level are used.
These can be seen in the model script which is available
[online](https://github.com/npct/pct/blob/master/models/aggregate-model.R).

```{r, echo=FALSE, fig.cap="National vs local cycling characteristics with hilliness, captured in the model results"}
source("models/aggregate-model.R") # this model creates the variable 'slc'
cormod <- cor(flow$clc, mod_logsqr$fitted.values) # crude indication of goodness-of-fit
# summary(mod_logsqr)

mod_nat <- readRDS("pct-bigdata/national/mod_logsqr_national_8.Rds")
justdist <- data.frame(
  dist_fast = 1:20,
  avslope = 0,
  type = "Flat"
)

justdist2 <- data.frame(
  dist_fast = 1:20,
  avslope = 2,
  type = "Hilly"
)

justdist3 <- data.frame(
  dist_fast = 1:20,
  avslope = 0,
  type = "Flat"
)

justdist4 <- data.frame(
  dist_fast = 1:20,
  avslope = 2,
  type = "Hilly"
)

justdist5 <- rbind(justdist3, justdist4)
justdist5$model <- "Local"

justdist <- rbind(justdist, justdist2)
justdist$model <- "National"

justdist$npred <- exp(predict(mod_nat, justdist))
justdist5$npred <- exp(predict(mod_logsqr, justdist))

justdist <- rbind(justdist, justdist5)

ggplot(justdist) +
  geom_line(aes(dist_fast, npred, color = model, linetype = type)) +
  xlab("Route distance (km)") + ylab("Expected number cycling")

dfcos <- round(rbind(coef(mod_nat), coef(mod_logsqr)), 3)
dfcos <- cbind(Model = c("National", "Local"), dfcos)
```

The correlation between fitted and observed cycling in the model is
`r round(cormod, 2)`, compared with 0.39 nationally.

The values for the coefficients are presented in the table below.

```{r, echo=FALSE}
library(knitr)
kable(dfcos, digits = 3)
```

## Cycling in the study area

```{r, echo=FALSE}
rcycle <- round(100 * sum(l$Bicycle) / sum(l$All), 1)
natcyc <- sum(luk$Bicycle) / sum(luk$All)
```

The overall rate of cycling in the flows in the study area
(after subsetting for distance) is `r rcycle`%, compared a
rate from the national data (of equally short flows)
of 5.0%.

## Scenarios

```{r, echo=FALSE, fig.cap="Rate of cycling in model scenarios. Note the total percentage cycling is equal to the area under each line."}

l$slc <- flow$plc
l$base_olc <- l$Bicycle
l$base_slc <- l$slc * l$All
l$base_sic <- l$base_slc - l$base_olc
# l$sic2 <- l$slc * l$All - l$Bicycle # identical sic result

# # # # # # # # # # # # #
# Additional scenarios  #
# # # # # # # # # # # # #

# Additional scenarios
# Replace with source("models/aggregate-model-dutch|gendereq|ebike.R"))
set.seed(2015)
l$npred <- exp(predict(mod_nat, flow))

l$cdp_slc <-l$All * (l$clc + l$npred) # may be more that 1
l$cdp_sic <- l$cdp_slc - l$Bicycle

l$gendereq_slc <- l$All * (l$slc + runif(nrow(l), 0, max = 0.1))
l$gendereq_sic <- l$gendereq_slc - l$base_olc

l$dutch_slc <- l$All * (l$slc + runif(nrow(l), 0, max = 0.2))
l$dutch_sic <- l$dutch_slc - l$base_olc

l$ebike_slc <- l$All * (l$slc + runif(nrow(l), 0, max = 0.3))
l$ebike_sic <- l$ebike_slc - l$base_olc

dfscen <- dplyr::select(l@data, contains("slc"), All, olc = Bicycle, dist_fast)
dfscen <- dfscen[-which(names(dfscen) == "slc")]
dfscen <- dfscen[-which(names(dfscen) == "base_slc")]
# head(dfscen)

library(tidyr)
dfsp <- gather(dfscen, key = scenario, value = slc, -dist_fast)
# head(dfsp)
dfsp$scenario <- factor(dfsp$scenario)
 summary(dfsp$scenario)
dfsp$scenario <- 
  factor(dfsp$scenario, levels = levels(dfsp$scenario)[c(5, 4, 1, 2, 3, 6)])
scalenum <- sum(l$All) 

ggplot(dfsp) +
  geom_freqpoly(aes(dist_fast, weight = slc,
    color = scenario), binwidth = 1) + 
  ylab("Total number of trips") +
  xlim(c(0,12))

dfsp$dist_band <- cut(dfsp$dist_fast, c(0, 2, 5, 10, 20))
dfsum <- summarise(group_by(dfsp, scenario, dist_band), Percent = sum(slc) / sum(l$All))
dfsum$Percent <- dfsum$Percent 
dfspread <- spread(dfsum, scenario, Percent)
dfspread$dist_band <- as.character(dfspread$dist_band)
dfspreadf <-c("Total", colSums(dfspread[2:7]))
dfspread[3:7] <- do.call(cbind, apply(dfspread[3:7], 2, function(x) round(x / dfspread[2] * 100, 2)))
dfspread <- rbind(dfspread, dfspreadf)
dfspread <- dfspread[c(1, 2, 7, 3, 4, 5, 6)]
```

```{r, echo=FALSE}
kable(dfspread)
```


```{r}
# # # # # # # # # # # # # # # # # #
# Extract area-level commute data #
# # # # # # # # # # # # # # # # # #

for(i in 1:nrow(cents)){

  # all flows originating from centroid i
  j <- which(l$Area.of.residence == cents$geo_code[i])

  cents$base_olc[i] <- sum(l$Bicycle[j])
  cents$base_slc[i] <- sum(l$base_slc[j])
  cents$base_sic[i] <- sum(l$base_sic[j])

  # values for scenarios
  cents$cdp_slc[i] <- sum(l$cdp_slc[j])
  cents$cdp_sic[i] <- sum(l$cdp_sic[j])
  
  cents$gendereq_slc[i] <- sum(l$gendereq_slc[j])
  cents$gendereq_sic[i] <- sum(l$gendereq_sic[j])

  cents$dutch_slc[i] <- sum(l$dutch_slc[j])
  cents$dutch_sic[i] <- sum(l$dutch_sic[j])

  cents$ebike_slc[i] <- sum(l$ebike_slc[j])
  cents$ebike_sic[i] <- sum(l$ebike_sic[j])

  cents$av_distance[i] <- sum(l$dist[j] * l$All[j])  / sum(l$All[j])
  cents$cirquity[i] <- sum(l$cirquity[j] * l$All[j], na.rm = T )  / sum(l$All[j])
  cents$distq_f[i] <- sum(l$distq_f[j] * l$All[j], na.rm = T )  / sum(l$All[j])
}

names(l) # which line names can be added for non-directional flows?
# dput(c(names(l)[addids], c("cdp_slc", "cdp_sic")))
addidsn <- c("All", "Work.mainly.at.or.from.home", "Underground..metro..light.rail..tram", 
"Train", "Bus..minibus.or.coach", "Taxi", "Motorcycle..scooter.or.moped", 
"Driving.a.car.or.van", "Passenger.in.a.car.or.van", "Bicycle", 
"On.foot", "Other.method.of.travel.to.work", "base_olc", "base_slc", 
"base_sic", "gendereq_slc", "gendereq_sic", "dutch_slc", "dutch_sic", 
"ebike_slc", "ebike_sic", "cdp_slc", "cdp_sic")
addids <- which(names(l) %in% addidsn)
# addids <- c(3:14, 23:31)
# summary(l[addids])


# Aggregate bi-directional flows

# Subset by zone bounding box
# l <- l[as.logical(gContains(zone, l, byid = T)),]
nrow(l)

# 4: by aggregating 2 way flows
l <- gOnewayid(l, attrib = c(addids))

l$clc <- l$Bicycle / l$All
l$slc <- l$base_slc / l$All

nrow(l)
idsel <- l$id
plot(zbuf)
lines(l, col = "green")
rf <- rf[rf@data$id %in% idsel,]
rq <- rq[rq@data$id %in% idsel,]

# Sanity test
summary(l@data)
cents_ttwa <- cents # copy cents data (we'll overwrite cents)

# # Subset to zone
# cents <- cents_ttwa[zone,] # subset centroids geographically
# zones <- zones[cents,]
zbuf <- spTransform(zbuf, CRS("+init=epsg:4326"))
plot(zbuf)
plot(zones, add = T)
points(cents_ttwa, col = "red")
lines(l, col = "black")
lines(rq, col = "green")
lines(rf, col = "blue")
```



```{r, echo=FALSE}
# # # # # # # # #
# Save the data #
# # # # # # # # #

# Transfer cents data to zones
c_in_z <- names(cents) == "avslope"
zones@data <- left_join(zones@data, cents@data[,!c_in_z])
summary(cents)
summary(zones)

# 
# # Save objects
# saveRDS(zones, paste0("pct-data/", la, "/z.Rds"))
# saveRDS(cents, paste0("pct-data/", la, "/c.Rds"))
# saveRDS(l, paste0("pct-data/", la, "/l.Rds"))
# saveRDS(rf, paste0("pct-data/", la, "/rf.Rds"))
# saveRDS(rq, paste0("pct-data/", la, "/rq.Rds"))
# saveRDS(mod_logsqr, paste0("pct-data/", la, "/model.Rds"))
# 
# # # Save data for wider ttwz area
# # saveRDS(ttwa_zone, paste0("pct-data/", la, "/ttw_zone.Rds"))
# # saveRDS(cents_ttwa, paste0("pct-data/", la, "/c_ttwa.Rds"))
# # saveRDS(l_ttwa, paste0("pct-data/", la, "/l_ttwa.Rds"))
# 
# # Create new folder in pct-shiny repo
# rname <- tolower(la)
# dname <- paste0("~/repos/pct-shiny/", rname, "/")
# dir.create(dname)
# files <- list.files("~/repos/pct-shiny/manchester/", full.names = T)
# file.copy(files, dname)
# server <- readLines(paste0(dname, "server.R"))
# server <- gsub("manchester", la, server)
# writeLines(server, paste0(dname, "server.R"))
# 
# # Save the script that loaded the lines into the data directory
# file.copy("loading-data/load.R", paste0("pct-data/", la, "/load.R"))
```

## Time taken

The time taken to run the analysis for this area is presented below.

```{r}
end_time <- Sys.time()

end_time - start_time
```


