---
title: "National Propensity to Cycle Tool - local results"
author: "Created by the NPCT team"
output:
  html_document:
    fig_caption: yes
    highlight: pygments
    theme: null
    toc: yes
---

```{r, include=FALSE}
# TODO: Fix buildmaster so it works for Devon AND Norfolk
start_time <- Sys.time() # for timing the script
source("set-up.R") # pull in packages needed
```

This document was produced automatically at `r start_time`.

## Introduction

The results of the National Propensity to Cycle Tool (NPCT) scenarios are based on a model.
This document presents information about the input data, model diagnostics,
run time and the key outputs for each region.
The aim is to inform users of the NPCT's interactive map how the results were generated.
This document assumes some technical knowledge in the area of transport planning.

The code underlying the NPCT is open source, under the conditions of the MIT License.
Both the [pct](https://github.com/npct/pct) and [pct-shiny](https://github.com/npct/pct-shiny) and be modified by others provided attribution to the original.

## Initial parameters

The preset values used to select the study area and filter the origin destination data are described in this section.

```{r, echo=FALSE}
# Create default LA name if none exists
if(!exists("region")) region <- "Avon" 
```


```{r, warning=FALSE}
# Set local authority and ttwa zone names
region # name of the region
region_path <- file.path("pct-data", region)
if(!dir.exists(region_path)) dir.create(region_path) # create data directory

# Minimum flow between od pairs, subsetting lines. High means fewer lines.
mflow <- 100
mdist <- 15 # maximum euclidean distance (km) for subsetting lines
min_zones <- 60 # minumum number of zones in study area before buffer used
buff_dist <- 0 # buffer (km) used to select additional zones (often zero = ok)
```

## Input zone data

The input zones area are summarised in this section.

```{r plotzones, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
# ukmsoas <- shapefile("pct-bigdata/national/msoas.shp")
# gMapshape("pct-bigdata/national/msoas.shp", 8)
# ukmsoas <- shapefile("pct-bigdata/national/msoasmapshaped_8%.shp")
# ukmsoas <- spTransform(ukmsoas, CRS("+init=epsg:4326"))
# geojson_write(ukmsoas, file = "pct-bigdata/national/msoas.geojson")
ukmsoas <- geojson_read("pct-bigdata/national/msoas.geojson", what = "sp")

# Load population-weighted centroids
cents <- readOGR("pct-bigdata/national/cents.geojson", layer = "OGRGeoJSON")
cents$geo_code <- as.character(cents$geo_code)

# Load local authorities and districts
las <- readOGR(dsn = "pct-bigdata/national/las-pcycle.geojson", layer = "OGRGeoJSON")
# las <- readOGR(dsn = "pct-bigdata/national/cuas-mf.geojson", layer = "OGRGeoJSON")
regions <- geojson_read("pct-bigdata/national/regions.geojson", what = "sp")
las_cents <- SpatialPoints(coordinates(las))
las_in_region <- rgeos::gIntersects(las_cents, region_shape, byid = T)
las_in_region <- las_in_region[1,]
summary(las_in_region)
las_in_region <- las[las_in_region,]

proj4string(cents) <- proj4string(regions) # assign correct CRS
sel_region <- grepl(pattern = region, x = regions$Region, ignore.case = T)
```

The selected region was `r as.character(regions$Region[sel_region ])`.


```{r, include=FALSE}
# lasdat <- SpatialPointsDataFrame(coords = coordinates(las), data = las@data)

# Load counties and unitary authorities
# cuas <- readOGR(dsn = "pct-bigdata/national/cuas.geojson", layer = "OGRGeoJSON")
# proj4string(lasdat) <- proj4string(cuas)
# cuas <- aggregate(lasdat, cuas, mean, na.action = na.omit()) # todo: fix data
# # tmap::qtm(cuas2,fill = "clc")

# create region shape (and add buffer in m)
region_shape <- region_orig <- 
  regions[grep(pattern = region, x = regions$Region, ignore.case = T),]
region_shape <- spTransform(region_shape, CRS("+init=epsg:27700"))
region_shape <- region_shape <- gBuffer(region_shape, width = buff_dist)
region_shape <- spTransform(region_shape, proj4string(cents))

# select msoas of interest
cents <- cents[region_shape,]
zones <- ukmsoas[ukmsoas@data$geo_code %in% cents$geo_code, ]
```

The characteristics of zones are as follows:

```{r, echo=FALSE}
nzones <- nrow(zones) # how many zones?
mzarea <- round(median(gArea(zones, byid = T) / 10000), 1) # average area of zones, sq km
```

- Number of zones: `r nzones`, compared with 6791 in England
- Median area of zones: `r mzarea` ha, compared with 300 ha across England

## Input flow data

```{r, echo=FALSE, results='hide'}
# flow <- readRDS("pct-bigdata/national/flow_eng_avlslope.Rds")
flow <- readRDS("private-data/flowsex-merged.Rds")


# Subset by zones in the study area
o <- flow$Area.of.residence %in% cents$geo_code
d <- flow$Area.of.workplace %in% cents$geo_code
flow <- flow[o & d, ] # subset OD pairs with o and d in study area
sel <- flow$All > mflow # subset OD pairs by n. people using it
```

```{r distance-dist, echo=FALSE, fig.cap="The study region (thick black border), selected zones (grey), the administrative zone region (red line) and local authorities (blue line). The black straight green represent the most intensive commuting OD pairs.", echo=FALSE, message=FALSE, warning=FALSE}
flow <- flow[sel, ]

l <- od2line(flow = flow, zones = cents)

# Calculate line lengths
l <- spTransform(l, CRS("+init=epsg:27700"))
l$dist <- gLength(l, byid = T) / 1000 # Euclidean distance
dsel <- l$dist < mdist
l <- spTransform(l, proj4string(cents))

plot(zones, col = "lightgrey")
plot(regions, add = T)
plot(las_in_region, border = "blue", add = T, lwd = 2)
plot(region_orig, lwd = 5, add = T)
plot(region_shape, border = "red", add = T, lwd = 2)
lines(l[l$All > 100,], col = "green")
```

```{r, echo=FALSE}
l <- l[dsel,]
l <- l[l$dist > 0, ] # to remove OD pairs of 0 length
# l <- l[1:30,] # warning: subset lines arbitrarily
```

```{r, echo=FALSE}
# nrow(flow) # how many OD pairs in the study area?
# proportion of OD pairs in min-flow based subset
pmflow <- round(sum(sel) / nrow(flow) * 100, 1)
# % all trips covered
pmflowa <- round(sum(flow$All[sel]) / sum(flow$All) * 100, 1)
```

There are **`r nrow(flow)`** OD pairs with origins and destinations in the study
area. Of these, **`r sum(sel)`** meet the criteria that at least `r mflow` people 
reported travelling between these two areas in the 2011 Census.
That is  **`r pmflow`%** of OD pairs, accounting for
**`r pmflowa`%** of commutes in the study area.

## Hilliness of OD pairs

The average hilliness of zones in the study area is
`r round(mean(zones$avslope), 1)`
degrees,
compared with the national average of 
`r round(mean(ukmsoas$avslope, na.rm = T), 1)`. This data is displayed in the
figure below.

```{r, echo = FALSE}
tm_shape(zones) +
  tm_fill("avslope", n = 3, palette = "Oranges")
```

```{r, echo=FALSE}
# Hilliness of OD pairs
# (calculated as the average gradient of the zone
# of the flow's origin and destination, in degrees)
# is 
# `r round(mean(flow$avslope * flow$All / mean(flow$All), na.rm = T), 2)`.
# The UK
# average is xx degrees
```

## Lines allocated to the road network

We use CycleStreets.net to estimate optimal routes. 
An illustration of these routes is presented below.

```{r flow-vars, echo=FALSE}
# # # # # # # # # # # # # # # # # #
# Calculate flow-level variables: #
# distances and olc for ag. model #
# # # # # # # # # # # # # # # # # #

# Calculate distances (eventually use route distance)

# # # # # # # # # # # # # # #
# Allocate OD pairs to network #
# Warning: time-consuming!  #
# Needs CycleStreet.net API #
# # # # # # # # # # # # # # #


f <- list.files(paste0("pct-data/", region, "/"))
if(sum(grepl("l", f)) == 0){
  saveRDS(l, paste0("pct-data/", region, "/l_all.Rds"))
}

# Create or load route allocated lines (depending on past loading)
if(file.exists(paste0("pct-data/", region, "/rf_ttwa.Rds"))){
  rf <- readRDS(paste0("pct-data/", region, "/rf_ttwa.Rds")) # if you've loaded them
  rq <- readRDS(paste0("pct-data/", region, "/rq_ttwa.Rds"))
  # l <- readRDS(paste0("pct-data/", region, "/l.Rds"))
  # nrow(rf) == nrow(l)
} else {
  # see ?line2route from stplanr package
  # mytoken <- "f3fe3d078ac34737" # warning: not a real key 
  mytoken <- readLines("~/Dropbox/dotfiles/cyclestreets-api-key-rl")
  Sys.setenv(CYCLESTREET = mytoken)
  rf <- line2route(l[ l$dist > 0, ], silent = T)
  rq <- line2route(l[ l$dist > 0, ], plan = "quietest", silent = T)
}
saveRDS(rf, paste0("pct-data/", region, "/rf_ttwa.Rds")) # save the routes
saveRDS(rq, paste0("pct-data/", region, "/rq_ttwa.Rds"))
  
path_fail <- !1:nrow(l) %in% as.numeric(row.names(rf)) | 
  !1:nrow(l) %in% as.numeric(row.names(rq))
  
l <- l[!path_fail,]

# Process route data
rf$length <- rf$length / 1000
rq$length <- rq$length / 1000
rf$id <- l$id
rq$id <- l$id
proj4string(rf) <- proj4string(l)
proj4string(rq) <- proj4string(l)
```

```{r plot-rlines, warning=FALSE, echo = FALSE, fig.cap="Straight and route-lines allocated to the travel network"}
library(sp)
plot(l)
lines(rf, col = "red")
lines(rq, col = "green")
```

Of the lines allocated to the route network, `length(!path_fail)` failed.

```{r, echo=FALSE}

# Allocate route factors to OD pairs
# nz <- which(l$dist > 0) # non-zero lengths = nz
l$dist_quiet <- l$dist_fast <- l$cirquity <- l$distq_f <- NA
if(!nrow(rf) == nrow(l)) print("Warning, lines and routes are different lengths")
  l$dist_fast <- rf$length
  l$dist_quiet <- rq$length
  l$cirquity <- rf$length / l$dist
  l$distq_f <- rq$length / rf$length
```

## Distance distributions

The distance distribution of trips in the study area is displayed in the figure below, which compares the result with the distribution of trips nationwide.

```{r, echo=FALSE, fig.cap="Distance distribution of all trips in study lines (blue) compared with national average (dotted bars)"}
luk <- readRDS("pct-bigdata/national/l_sam8.Rds")

hdfl <- dplyr::select(l@data, All, dist_fast)
hdfl$Scope <- "Local"
hdfl$All <- hdfl$All / sum(hdfl$All)

hdfu <- dplyr::select(luk@data, All, dist_fast)
hdfu$Scope <- "National"
hdfu$All <- hdfu$All / sum(hdfu$All)

histdf <- rbind(hdfl, hdfu)

ggplot(histdf) +
  geom_histogram(aes(dist_fast, weight = All, fill = Scope, linetype = Scope),
    position = "identity", colour = "black", binwidth = 0.5) +
  scale_fill_manual(values = c("lightblue", NA)) +
  scale_linetype(c(1, 2), guide = "none") +
  scale_y_continuous() + 
  # scale_y_continuous(labels = percent) + 
  xlab("Route distance (km)") + 
  ylab("Proportion of trips in each band") + 
  xlim(c(0,13)) + 
  theme_bw()

pl5kmuk <- round(sum(luk$All[luk$dist_fast < 5]) /
    sum(luk$All) * 100, 1)
pl5km <- round(sum(l$All[l$dist_fast < 5]) /
    sum(l$All) * 100, 1)
```

From the nationwide sample of trips, `r pl5kmuk`% of trips are less than 5km.

In the case study area
`r pl5km`% of sampled trips are less than 5km.

Subsetting by distance (set
to `r mdist` km) and removing inter-zone OD pairs
further reduces the number of OD pairs from `r sum(sel)`
to `r nrow(l)`.

```{r, echo=FALSE}
# a = 11
# plot(l[a,])
# lines(rf[a,], col = "red")
# lines(rq[a,], col = "green")

# # # # # # # # # # # # # #
# Estimates slc from olc  #
# # # # # # # # # # # # # #

l$clc <- l$Bicycle / l$All
flow_ttwa <- flow # save OD pairs for the ttwa
flow <- l@data
```

## The flow model

To estimate the potential rate of cycling under different scenarios
regression models operating at the flow level are used.
These can be seen in the model script which is available
[online](https://github.com/npct/pct/blob/master/models/aggregate-model.R).

```{r, echo=FALSE}
source("models/aggregate-model.R") # this model creates the variable 'slc'
cormod <- cor(flow$clc, mod_logsqr$fitted.values) # crude indication of goodness-of-fit
# summary(mod_logsqr)

mod_nat <- readRDS("pct-bigdata/national/mod_logsqr_national_8.Rds")
```

## Cycling in the study area

```{r, echo=FALSE}
rcycle <- round(100 * sum(l$Bicycle) / sum(l$All), 1)
natcyc <- sum(luk$Bicycle) / sum(luk$All)
```

The overall rate of cycling in the OD pairs in the study area
(after subsetting for distance) is `r rcycle`%, compared a
rate from the national data (of equally short OD pairs)
of 5.0%.

## Scenarios

```{r, include=FALSE}
l$slc <- flow$plc
l$base_olc <- l$Bicycle
l$base_slc <- l$slc * l$All
l$base_sic <- l$base_slc - l$base_olc
# l$sic2 <- l$slc * l$All - l$Bicycle # identical sic result

# # # # # # # # # # # # #
# Additional scenarios  #
# # # # # # # # # # # # #

# Additional scenarios
# Replace with source("models/aggregate-model-dutch|gendereq|ebike.R"))
set.seed(2015)
l$npred <- exp(predict(mod_nat, flow))

l$cdp_slc <-l$All * (l$clc + l$npred) # may be more that 1
l$cdp_sic <- l$cdp_slc - l$Bicycle

# gendereq scenario
p_trips_male <- mean(las_in_region$clc_m) # proportion of bicycle trips by males

clc_m <- l$Bicycle * p_trips_male
pmale_c <- clc_m / l$Male
slc_gendereq_f <- l$Female * pmale_c
slc_gendereq <- clc_m + slc_gendereq_f

l$gendereq_slc <- slc_gendereq
l$gendereq_sic <- l$gendereq_slc - l$base_olc

# Dutch scenario - coefficients calculated from Dutch NTS by A. Goodman
mod_dutch <- mod_nat
mod_dutch$coefficients[1] <- -0.3253
mod_dutch$coefficients[2] <- -0.3543
mod_dutch$coefficients[3] <- 0.7750
l$dutch_slc <- l$All * exp(predict(mod_dutch, flow))
l$dutch_sic <- l$dutch_slc - l$base_olc

mod_ebike <- mod_dutch
mod_ebike$coefficients[2] <- -0.3

l$ebike_slc <- l$All * exp(predict(mod_ebike, flow))
l$ebike_sic <- l$ebike_slc - l$base_olc

dfscen <- dplyr::select(l@data, contains("slc"), All, olc = Bicycle, dist_fast)
dfscen <- dfscen[-which(names(dfscen) == "slc")]
dfscen <- dfscen[-which(names(dfscen) == "base_slc")]
# head(dfscen)

dfsp <- gather(dfscen, key = scenario, value = slc, -dist_fast)
# head(dfsp)
dfsp$scenario <- factor(dfsp$scenario)
 summary(dfsp$scenario)
dfsp$scenario <- 
  factor(dfsp$scenario, levels = levels(dfsp$scenario)[c(5, 4, 1, 2, 3, 6)])
scalenum <- sum(l$All) 
```

```{r, echo=FALSE, warning=FALSE, fig.cap="Rate of cycling in model scenarios. Note the total percentage cycling is equal to the area under each line."}
ggplot(dfsp) +
  geom_freqpoly(aes(dist_fast, weight = slc,
    color = scenario), binwidth = 1) + 
  ylab("Total number of trips") +
  xlab("Route distance (km)") +
  scale_color_discrete(name = "Mode and\nscenario\n(cycling)") +
  xlim(c(0,12)) +
  theme_bw()

dfsp$dist_band <- cut(dfsp$dist_fast, c(0, 2, 5, 10, 20))
dfsum <- summarise(group_by(dfsp, scenario, dist_band), Percent = sum(slc) / sum(l$All))
dfsum$Percent <- dfsum$Percent 
dfspread <- spread(dfsum, scenario, Percent)
dfspread$dist_band <- as.character(dfspread$dist_band)
dfspreadf <- c("Total", round(colSums(dfspread[2:7])* 100, 1))
dfspread[3:7] <- do.call(cbind, apply(dfspread[3:7], 2, function(x) round(x / dfspread[2] * 100, 1)))
dfspread <- rbind(dfspread, dfspreadf)
dfspread <- dfspread[c(1, 2, 7, 3, 4, 5, 6)]
dfspread$All <- round(as.numeric(dfspread$All) * 100, 1)
dfspread$All[nrow(dfspread)] <- dfspread$All[nrow(dfspread)] / 100
```

The table below illustrates the same information by distance band.

```{r, echo=FALSE}
# names(dfspread)[1:3] <- c("Distance band", 	"All modes", "Observed level (OLC)")
kable(dfspread, format = "html", digits = 1, )
```


```{r, include=FALSE}
# # # # # # # # # # # # # # # # # #
# Extract area-level commute data #
# # # # # # # # # # # # # # # # # #

for(i in 1:nrow(cents)){

  # all OD pairs originating from centroid i
  j <- which(l$Area.of.residence == cents$geo_code[i])

  cents$base_olc[i] <- sum(l$Bicycle[j])
  cents$base_slc[i] <- sum(l$base_slc[j])
  cents$base_sic[i] <- sum(l$base_sic[j])

  # values for scenarios
  cents$cdp_slc[i] <- sum(l$cdp_slc[j])
  cents$cdp_sic[i] <- sum(l$cdp_sic[j])
  
  cents$gendereq_slc[i] <- sum(l$gendereq_slc[j])
  cents$gendereq_sic[i] <- sum(l$gendereq_sic[j])

  cents$dutch_slc[i] <- sum(l$dutch_slc[j])
  cents$dutch_sic[i] <- sum(l$dutch_sic[j])

  cents$ebike_slc[i] <- sum(l$ebike_slc[j])
  cents$ebike_sic[i] <- sum(l$ebike_sic[j])

  cents$av_distance[i] <- sum(l$dist[j] * l$All[j])  / sum(l$All[j])
  cents$cirquity[i] <- sum(l$cirquity[j] * l$All[j], na.rm = T )  / sum(l$All[j])
  cents$distq_f[i] <- sum(l$distq_f[j] * l$All[j], na.rm = T )  / sum(l$All[j])
}

# names(l) # which line names can be added for non-directional OD pairs?
# dput(c(names(l)[addids], c("cdp_slc", "cdp_sic")))
addidsn <- c("All", "Work.mainly.at.or.from.home", "Underground..metro..light.rail..tram", 
"Train", "Bus..minibus.or.coach", "Taxi", "Motorcycle..scooter.or.moped", 
"Driving.a.car.or.van", "Passenger.in.a.car.or.van", "Bicycle", 
"On.foot", "Other.method.of.travel.to.work", "base_olc", "base_slc", 
"base_sic", "gendereq_slc", "gendereq_sic", "dutch_slc", "dutch_sic", 
"ebike_slc", "ebike_sic", "cdp_slc", "cdp_sic")
addids <- which(names(l) %in% addidsn)
# addids <- c(3:14, 23:31)
# summary(l[addids])


# Aggregate bi-directional OD pairs

# Subset by zone bounding box
# l <- l[as.logical(gContains(zone, l, byid = T)),]
# nrow(l)

# 4: by aggregating 2 way OD pairs
l <- gOnewayid(l, attrib = c(addids))

l$clc <- l$Bicycle / l$All
l$slc <- l$base_slc / l$All

nrow(l)
idsel <- l$id
rf <- rf[rf@data$id %in% idsel,]
rq <- rq[rq@data$id %in% idsel,]

# Sanity test
head(l@data[1:5])
cents_ttwa <- cents # copy cents data (we'll overwrite cents)

# # Subset to zone
# cents <- cents_ttwa[zone,] # subset centroids geographically
# zones <- zones[cents,]
```

```{r, echo=FALSE, results='hide', fig.cap="Illustration of OD pairs on travel network"}
zbuf <- spTransform(zbuf, CRS("+init=epsg:4326"))
plot(zbuf)
plot(zones, add = T)
points(cents_ttwa, col = "red")
lines(l, col = "black")
lines(rq, col = "green")
lines(rf, col = "blue")
```

## Flow model results

To estimate the potential rate of cycling under different scenarios
regression models operating at the flow level are used.
These can be seen in the model script which is available
[online](https://github.com/npct/pct/blob/master/models/aggregate-model.R).

```{r, echo=FALSE, fig.cap="National vs local cycling characteristics with hilliness, captured in the model results"}
source("models/aggregate-model.R") # this model creates the variable 'slc'
cormod <- cor(flow$clc, mod_logsqr$fitted.values) # crude indication of goodness-of-fit
# summary(mod_logsqr)

mod_nat <- readRDS("pct-bigdata/national/mod_logsqr_national_8.Rds")

justdist1 <- data.frame(
  dist_fast = 1:20,
  avslope = 1,
  type = "Flat")

justdist2 <- justdist1
justdist2$avslope <- 1.5
justdist2$type <- "Hilly"

justdist <- rbind(justdist1, justdist2) # for prediction
justdist$model <- "National"

justdist5 <- justdist6 <- justdist7 <- justdist # replicate
justdist5$model <- "Local"
justdist6$model <- "Dutch"
justdist7$model <- "Ebike"

justdist$npred <- exp(predict(mod_nat, justdist))
justdist5$npred <- exp(predict(mod_logsqr, justdist))
justdist6$npred <- exp(predict(mod_dutch, justdist))
justdist7$npred <- exp(predict(mod_ebike, justdist))

justdist <- rbind(justdist, justdist5, justdist6, justdist7)

ggplot(justdist) +
  geom_line(aes(dist_fast, npred, color = model, linetype = type),
    size = 1.5) +
  xlab("Route distance (km)") + ylab("Expected proportion cycling") +
  theme_bw()

dfcos <- round(rbind(coef(mod_nat), coef(mod_logsqr)), 3)
dfcos <- cbind(Model = c("National", "Local"), dfcos)
```

The correlation between fitted and observed cycling in the model is
`r round(cormod, 2)`, compared with 0.39 nationally.

The values for the coefficients are presented in the table below.

```{r, echo=FALSE}
dfcos <- data.frame(dfcos)
names(dfcos) <- c("Model", "Alpha", "Distance", "Dist^0.5", "Hills", "Dist/Hills")
kable(dfcos, digits = 3)
```

## Network analysis

Now we aggregate the overlapping routes to create a route network.  
The value of each segment in the network corresponds to the total number of cyclists who we estimate to use the segment.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="The route network, with widths proportional to the current estimated number of commuter cyclists"}
# Scenario names - may need updating
# nrow(l) == nrow(rf)
rf@data["base_olc"] <- l@data["base_olc"]
rnet <- gOverline(rf, "base_olc")
rf$clc <- NULL
# test the resulting plot
plot(rnet, lwd = rnet$base_olc / mean(rnet$base_olc))
scens <- c("cdp_slc", "gendereq_slc", "dutch_slc", "ebike_slc")
for(i in scens){
  rf@data[i] <- l@data[i]
  rnet_tmp <- gOverline(rf, i)
  rnet@data[i] <- rnet_tmp@data[i]
  rf@data[i] <- NULL
}
# plot(rnet, lwd = rnet$cdp_slc / mean(rnet$base_olc))
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# # # # # # # # #
# Save the data #
# # # # # # # # #

# Rename columns
l@data <- rename(l@data, Home = Area.of.residence, Work = Area.of.workplace)
names(zones@data)


# Transfer cents data to zones
c_in_z <- names(cents) == "avslope"
zones@data <- left_join(zones@data, cents@data[,!c_in_z])
# summary(cents)
# summary(zones)

# 
# # Save objects
# Save objects # uncomment these lines to save model output
saveRDS(zones, paste0("pct-data/", region, "/z.Rds"))
saveRDS(cents, paste0("pct-data/", region, "/c.Rds"))
saveRDS(l, paste0("pct-data/", region, "/l.Rds"))
saveRDS(rf, paste0("pct-data/", region, "/rf.Rds"))
saveRDS(rq, paste0("pct-data/", region, "/rq.Rds"))
saveRDS(rnet, paste0("pct-data/", region, "/rnet.Rds"))
saveRDS(mod_logsqr, paste0("pct-data/", region, "/model.Rds"))

# 
# # Save the script that loaded the lines into the data directory
file.copy("load.Rmd", paste0("pct-data/", region, "/load.Rmd"))

# Create folder in shiny app folder (assuming it's next to pct)
mess <- paste0("cp -r ../pct-shiny/coventry/ ../pct-shiny/", region)
system(mess)
server_text <- readLines("../pct-shiny/coventry/server.R")
server_text <- gsub(pattern = "coventry", replacement = region, x = server_text)
writeLines(server_text, con = paste0("../pct-shiny/", region, "/server.R"))
```

 ## Time taken

The time taken to run the analysis for this area is presented below.

```{r}
end_time <- Sys.time()

end_time - start_time
```


