---
title: "Regression to find distance decay parameters"
author: "Robin Lovelace"
date: "01/27/2015"
output: pdf_document
---

The aim of this vignette is to demonstrate how the parameters of distance decay
can be estimated using a regression model on census flow data on travel to work.

The first thing is to load the data:

```{r, message=FALSE}
if( grepl("vig", getwd())) setwd("../")
source("set-up.R")
# load some flow data
flow <- read.csv("pct-data/leeds/sample-leeds-centre-dists.csv")
# load the zones
leeds <- readOGR("pct-data/leeds/", "leeds-central-sample")
```

Let's see what the flow data contain:

```{r}
head(flow)[c(1,2,3,4,10,14)]
```

Next step: calculate the proportion who cycle:

```{r}
flow$pcycle <- flow$Bicycle / flow$All.categories..Method.of.travel.to.work
```

Now lets find the determinants of propensity to cycle with linear regression.
First we linearize the equation: 

$$
p = \alpha * exp(-\beta d) (log both sides)
$$

$$
log(p) = -\beta d + log(\alpha)
$$

Now we can just run a linear model on the log of p, after subsetting to only include
distance values above 1 and adjusting p = 0 to be p = 0.001

```{r}
summary(flow$dist)
flow <- flow[ !flow$dist <= 0, ]
flow  <- flow[!flow$pcycle == 0, ] 
flow$dist <- flow$dist / 1000
plot(flow$dist, log(flow$pcycle))
plot(flow$dist, flow$pcycle)
```

Now we can run a linear models on this

```{r}
model1 <- lm(log(pcycle) ~ dist, data = flow ) 
summary(model1)
```

Next we can look at the dependency of these values on socio-demographic groups...

But to view the results of the model, let's make some graphs:

```{r}
alpha <- exp(model1$coefficients[1])
beta <- -model1$coefficients[2]
pred  <- alpha * exp(-beta * flow$dist )
hist(flow$pcycle, xlab = "Distance (km)")
hist(exp(pred))
```






