---
title: "The gender equality scenario"
author: "Robin Lovelace"
date: "May 2, 2015"
output: pdf_document
---

## Gender equality

The Gender Equality scenario (*gendereq*) is a relatively simple modification of
an existing scenario. For the purposes of explanation, we will
use the *observed level of cycling* (*OLC*) as the basis of the
scenario. However, it is possible to apply the *gendereq* method to
any scenario, as described towards the end of this section.

While the Baseline and Cycling Delivery Plan scenarios
(*base* and *cpd* respectively) assume that the local
gender split in cycling continues (typically, around 3/4 of cycle commuters
are males, although this varies over geographic space), *gendereq*
assumes that females become as likely to cycle as males are, for a given
origin-destination pair (per *flow*).
A prerequisite is a model-based estimate of the number of male
and female cyclists between origin and destinations for the baseline scenario.
This involves splitting the number of cyclists project by the model, 
the *Scenario-based Level of Cycling*, into
male ($SLC_m$) and female ($SLC_f$) components:

$$
SLC = SLC_m + SLC_f
$$

In the Gender Equality ($gendereq$) scenario
$SLC_m$ remains constant. $SLC_f$ changes, however, such that the proportion
of males cycling is equal to the proportion (or the probability)
of females cycling per flow
($pfemale(gendereq)_c = pmale_c$). 
Note that this is not as simple as setting $SLC_f = SLC_f$: the gender split
of the overall rate of flow must be taken into
account.^[To
illustrate this point, consider a flow in which there
are no female travellers. $SLC_f$ cannot possibly be positive in this case
because no females make the trip by any mode, so cannot switch to cycling.
]
Thus the *gendereq* scenario can be reduced to calculating $pcycle_m$:
the proportion of males for any give flow who make the trip by bicycle.
For each OD flow, we use the following variables to estimate the increase in female
propensity resulting from women being as likely to cycle as men per OD pair:

- The dominance of males amongst commuter cyclists per zone, expressed as an
estimate of the proportion of cycling by males ($pcycle(z)_m$).
- The male/female ratio of the travel-to-work flow in question, represented by
the proportion of all trips between origin and destination by males ($ptrips_m$).

Based on this data, the proportion of males per flow who cycle can be
calculated as follow:

$$
pmale_{c} = \frac{olc_m}{tflow * ptrips_m} 
$$

where

$$
olc_m = olc * pcycle_m
$$

and 

$$
pcycle_m = \frac{ptrips_m * pcycle(z)_m}{1 - ptrips_m * pcycle(z)_m  }
$$

To illustrate how this scenario works, imagine a scenario in which
5 from a total of 50 people commute by bicycle between zone i and j.
60% of the commutes overall between i and j are made by males and 75% of commuter
cycling in the zone is by males. This can be described in the language of R
as follows:

```{r}
tflow <- 50
olc <- 5 #  the currect rate of cycling between origin (o) and destination (d)
ptrips_m <- 0.6 # the proportion of all trips o and d by males
pcyclez_m <- 0.75 # the proportion of cycle trips in the zone/region made by males
pcycle_m <- ptrips_m * pcyclez_m / (1 - ptrips_m * pcyclez_m)
olc_m <- olc * pcycle_m
pmale_c <- olc_m / (tflow * ptrips_m)
```

We can verify this by calculating the number of female cyclists per flow
($olc_f$). In theory, $olc_m + olc_f$ should equal $olc$, set to 5 in this
hypothetical example:

```{r}
pcycle_f <- 1 - pcycle_m
olc_f <- olc * pcycle_f
pfemale_c <- olc_f / (tflow * (1 - ptrips_m))
print(pfemale_c)
print(olc_m)
print(olc_f)
olc_m + olc_f
```

SLC in the *gendereq* scenario ($SLC(gendereq)$) can then
be derived as follows:

$$
pfemale(gendereq)_c = pmale_c
$$

$$
slc(gendereq)_f = pfemale_c tflow_f
$$

Based on the above example, $SLC(gendereq)$
scenario would be 6.8 cyclists, up from 5 cyclists in OLC.
This represents an increase of just over 1/3 in the rate of cycling, for a 
reasonably typical example.

```{r}
pfemale_gendereq_c <- pmale_c
slc_gendereq_f <- pfemale_gendereq_c * (tflow * (1 - ptrips_m))
olc_m + slc_gendereq_f
```

Note that that $SLC(gendereq)$ can be calculated based on any scenario.
In our model, we estimate $SLC(gendereq)$ based on the Cycling Delivery Plan
($cdp$) scenario.