---
title: "National-level analysis of cycling potential"
author: "Robin Lovelace"
date: "July 5, 2015"
output:
  word_document:
    fig_caption: yes
  pdf_document: default
---


```{r, include=FALSE}
pkgs <- c("grid", "png", "knitr", "tmap", "dplyr", "ggmap")
lapply(pkgs, library, character.only = T)
opts_knit$set(root.dir = "../")
# source("analysis/la-analysis.R", print.eval = T)
```

```{r, echo=FALSE}
las <- readRDS("pct-bigdata/national/las.Rds")
# source("analysis/las-tables.R")
fixedBreaks = c(0,2,5,30)

t0 <- tail(las@data[order(las$pcycle),], 10)
t0 <- t0[10:1,]

t0 <- select(t0, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)

t0.1 <- head(las@data[order(las$pcycle),], 10)

t0.1 <- select(t0.1, Name = NAME, `Current %` = pcycle, `Expected %` = pcycle_exp, `Current (n)` = Bicycle_All, `Expected (n)` = expected, `% short car` = shortcar)

las$`Short trips (n)` <- las@data$All_.0.2. + las@data$All_.2.5.
las$`Short trips (%)` <- las$`Short trips (n)` / las@data$All_All * 100

```

The unit of analysis for the national-level analysis was the Local Authority District (LAD), of which there are 324 in England. This is
the highest geographical resolution for data that provides a breakdown
of mode of travel to work by distance. As with the flow-level model,
the analysis was undertaken using commuting data. This is the travel
dataset with the highest geographical resolution and greatest response
rate. Active travel patterns derived from commuting data at the this level also provides a reasonable
proxy for personal travel overall. We began the analysis be revisiting
the current rate of cycling, a vital starting point for understanding
the future cycling potential.

The spatial distribution of of commuter cycling across England
is depicted in Fig. 1. The dataset shows that there is wide variation
in the level of cycling, which can be split roughly into
3 categories. In 135 LADs (42% of the total number and population)
less than 2% of workers report cycling as their main means of travel
to work. Nearly half (160) of the 324 LADs have an intermediate level
of cycling, between 2 and 5%. 
The remaining 29 LADs have what can be considered a high rate of
cycling by English standards, of 5% of more. 
This means that only around 10% of the population can be considered
as living in areas where cycling is a normal mode of travel to work.
8 of the 'high cycling' LADs are
based in London.

Given the wide range of current levels of cycling across England,
it makes sense to allocate funding to each of the 3 types of
Local Authority. Authorities with a high current level of cycling
generally have a strong level of commitment to and community
support for cycling. New infrastructure along key 'cycling desire lines'
(e.g. as identified locally by the NPCT model) would likely
attract the 'near market' for cycling in these areas.
Areas with lower current levels of cycling will require greater
emphasis on complementary behavioural interventions in order to maximise the
value of infrastructure investement.



```{r, echo=FALSE, fig.cap="The current level of cycling to work in England from the 2011 Census."}
fixedBreaks = c(0,2,5,30)
qtm(las, "pcycle", fill.style = "fixed",
  fill.breaks = fixedBreaks, fill.title = "Percent cycling")
```

The next stage was to explore the correlates of cycling at the national
level. As expected, the proportion of short-distance trips (below 5 km)
was positively correlated with the rate of cycling (Fig. 2).
This analysis provides a simple yet effective way to identify
Local Authorities with a large unmet potential for cycling:
areas with an unexpectedly low relatively low rate of cycling
based on the high proportion of short-distance journey
could see large shifts to cycling with appropriate policy interventions.

The top 10 areas with the greatest unmet potential for cycling by
this measure are, in descending order,
Hastings, Blackburn with Darwen,
Burnley,                      Chesterfield,
Bradford,                     Oldham,
Hartlepool,                   Torbay,
Stoke-on-Trent  and             Calderdale.
Each of these local authorities are represented by a blue circle in
Fig. 2.

```{r, echo=FALSE, fig.cap="Scatterplot illustrating the link between the proportion of commutes that are short and the proportion of commutes that are cycled."}
lmod1 <- lm(pcycle ~ `Short trips (%)`, data = las@data)
# summary(lmod1)
sel <- c(which.min(lmod1$residuals), which.max(lmod1$residuals))
sel2 <- head(order(lmod1$residuals), 10)
sel3 <- tail(order(lmod1$residuals), 10)
# las$NAME[sel3]
# library(cowplot)
ggplot(las@data, aes(`Short trips (%)`, pcycle)) + geom_point() +
  geom_abline(intercept = -0.922, slope = 0.1092, col = "red") +
  geom_text(data = las@data[sel,], aes(`Short trips (%)`, pcycle,  label = NAME)) +
  geom_point(data = las@data[sel,], aes(`Short trips (%)`, pcycle), color = "red") +
  geom_point(data = las@data[sel2,], aes(`Short trips (%)`, pcycle), shape = 1, size = 5, col = "blue") +
  ylab("Level of cycling (%)")
  # geom_line(aes(lmod1$fitted.values, `Short trips (%)`), col = "red")
las$short_trip_residual <- lmod1$residuals
# write.csv(las@data, "pct-bigdata/national/la-ttw-distance.csv")
# This clear pattern The relationship between the average *length* of trips and the current ...
```

The analysis suggests that cycling interventions in
such areas (with more than around 40% of trips being less than 5 km
yet a low rate of cycling) could, if effective, yield a high return on investment. However, we note that the *cycling culture* of these areas is likely
to be weak and this may take time and 'soft' policy interventions,
in addition to high quality cycle infrastructure, to change.
Also, the example of South Cambridgeshire, in which only 21%
of commuting trips are 'short' yet the rate of cycling is high at
7.7%, demonstrates that
zones above the red line if Fig. 2 also have 
a high potential for cycling to grow.

```{r, echo=FALSE}
# Add % of 'cyclable' trips by bike
```


Applying the flow-level model to flows with distance distributions
representative of each authority allows for the scenarios described
in section xx to be applied nationwide. The results are presented
in Fig. 3 at the level of County and Unitary Authorities.
Note that 'cycling potential' can be measured in
absolute or proportional terms and that authorities with a higher
population tend to have a higher absolute potential increase in cycling.
The distinction is important when allocating funding.

```{r, echo=FALSE, fig.cap="Potential growth in cycling at the level of highway authorities in England", message=FALSE}
las$pcycle_cdp <- las$pcycle + las$pcycle_exp
las$Bicycle_exp <- las$All_All * las$pcycle_exp / 1000
fixedBreaks2 <- c(0,2,3,5)
fixedBreaks3 <- c(0,2,3,5)

library(rgdal)
cuas <- raster::shapefile("pct-bigdata/national/cuas.shp")
lasp <- rgeos::gCentroid(las, byid = T)
ldf <- las@data
output_data <- select(ldf, Bicycle_exp)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$Bicycle_exp <- cuasp@data$Bicycle_exp
qtm(cuas, "Bicycle_exp", fill.title = "Growth in commuter cyclists (n)")
# cuas@data$CTYUA12NM[order(cuas@data$Bicycle_exp, decreasing = T)]
output_data <- select(ldf, All_All)
lasp <- SpatialPointsDataFrame(lasp, output_data)
library(sp)
cuasp <- aggregate(lasp, cuas, FUN = sum)
# head(cuasp)
cuas$All <- cuasp@data$All_All
# summary(cuas$All)
# summary(head(cuas@data$All[order(cuas@data$Bicycle_exp, decreasing = T)], 10))
```

A question related to funding is the level at which cycling
investment will be allocated. If cycling policies are to form
part of an integrated national transport strategy, it is recommended
that funding for cycling is allocated at the level at which transport
investment for other modes takes place. Currently this is the level
of Combined and Unitary Authorities (CUAs) and, increasingly,
Combined Authorities and Local Enterprise Partnerships (LEPs).
CUAs with high unmet potential for cycling include
Lancashire, Hampshire and Kent. Our analysis suggests that
funding for cycling should move away from
sporadic and time-limited city-by-city investment to
a national-level strategy. Long-term funding for every Highway
Authority dedicated to cycling will enable a joined-up
and cost-effective cycling network to be created nationwide.
