---
title: "The flow model and user interface"
author: "Robin Lovelace"
date: "May 31, 2015"
output: 
  word_document: 
    fig_caption: yes
---

```{r, include=FALSE}
pkgs <- c("grid", "png", "knitr")
lapply(pkgs, library, character.only = TRUE)
opts_knit$set(root.dir = "../")
```


The main engine driving the local and national scenarios for the
NPCT in Phase I is a model operating at the flow level.
The unit of analysis is an origin-destination (OD) pair.
The approach was to use distance, hilliness and other
explanatory variables to predict distance decay
functions for the current rate of cycling. The geographical
distribution of cycling potential was estimated by applying
new distance decay curves to the existing transport flows,
allowing the NPCT to identify which 'desire lines' at the local
level have the greatest potential for increased cycling.

Overall, the model will identify routes in which there is a high
rate of travel over short distances yet a low rate of cycling.
Unexpectedly low rates of cycling, compared with model estimates,
are indicative of poor infrastructure, cultural aversion to cycling
or other factors not included in the model. The model does not
identify the reason for high or low cycling rates along particular
flow lines. However, it does draw attention to areas and routes
that have the greatest potential for increased rates of cycling
under various assumptions about the future.

The primary source of input data for the model is taken from the 
OD flow data results from the 2011 Census, provided by the WICID
site. `wu03ew_v2.csv` is a 104 MB (12 MB compressed)
comma-delimited file
of flows between unique origin destination pairs, disaggregated by mode.
The second key input is `cents.geojson`,
representing populations-weighted centroids of local
administrative zones (MSOAs).^[The license for these
datasets was accessed from http://www.nationalarchives.gov.uk/doc/open-government-licence/version/2/ in
March 2015.]

Details of the model are described in detail in a technical appendix.
Suffice to say that the model provides detailed break-downs of the
rate of cycling under each model scenario, taking into account the
current rate of travel, the cirquity of the local road network
and hilliness (more explanatory factors will be added in Phase II).
Example output from the model are presented in the figures below.

```{r, echo=FALSE, fig.cap="Prototype model output on the rate of cycling by distance band, and under a range of scenarios, for Manchester"}
# getwd()
grid.raster(readPNG("figures/mod-output-eg.png"))
```

```{r, echo=FALSE, fig.cap="Model output illustrating the geographic distribution of potential cycling uptake at the city level by area (blue:yellow) and for the 30 'desire lines' with the highest potential for cycling increases. The results represent an early implementation of the Cycling Delivery Plan (CDP) scenario in Coventry"}
grid.raster(readPNG("figures/mod-output-cov.png"))
```

The NPCT online tool was developed to allow for transport planners
to gain maximum benefit from the various outputs of the model results.
The primary interface to the model is through the 'Interactive map'
tab. This provides a range of options for interacting with
the model's outputs and for displaying
different types of model output via a number of widgets.
These widgets are both moveable and 'dockable', allowing the user
to focus most on the aspects of the map that are of most interest.
They are:

1. The 'Scenario' widget, a dropdown menu with cascading effects
affecting the model output for zones and lines.
2. The 'Attribute' dropdown menu, allowing users to focus
on the expected potential rate of cycling or increase current levels.
3. 'Cycling Flows', which allows the user to view key flows as
direct desire lines or as cyclist-optimised routes allocated
to the road network via the CycleStreets.net API.
4. The 'Freeze Lines' button, which allows the lines to move with 
the current map zoom or to stay fixed for zooming-in to the map.
5. 'Flows to show' allows the user to focus on specific set of
routes or take a more strategic view of the study area with
up to 50 lines (this can be increased in Phase II).

Additional controls are revealed with an additional button for
setting the basemap (allowing the user to quickly see the
current cycling network from Open Street Map) and displaying flow
data separately from zone data.

Additional tabs provide further information to users. These are:

- Lines Data and Zone Data, allowing users to see the raw
data underlying the model, focussing on the zones and lines
currently under investigation.
- Help, a tab providing instruction to new users of the tool. In 
Phase II of the project a video and interactive elements will
be added to assist with training and to make the tool more accessible.
- Model Output, a tab under development that will contain key
information about the study area (such as presented in the
Line Graph above). This will allow local transport planners to
compare their area quantitatively with England and other 
areas in terms of transport behaviour, rates of cycling
and potential growth rates. Critically, this final tab will
also present key model results such as the local parameters
determining distance decay the extent to which hilliness
seems to be a deterrent, compared with a model based on national
data.



