---
title: "Open online planning tools for sustainable transport"
author: "Robin Lovelace, University of Leeds"
date: "March 26, 2015"
output:
  ioslides_presentation:
    transition: slower
runtime: shiny
logo: ~/Dropbox/Public/img/logos/cdrc-square.png
---

```{r, echo=FALSE, include=FALSE}
pkgs <- c("grid", "png")
lapply(pkgs, library, character.only = T)
# bibliography: "~/Documents/R.bib"
```

## Presentation structure

> - The software (tools of the trade)
> - Context (motivations)
> - Building the National Propensity to Cycle Tool (NPCT)
> - Sustainable transport planning with R (stplanr)
> - Esperanzas y sueños

# I: The software

## Why R?

See the free, open source, online tutorial [github.com/robinlovelace/Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R).

```{r, echo=FALSE}
f <- "/home/robin/repos/Creating-maps-in-R/figure/facet_london.png"
grid.raster(readPNG(f))
```

## The packages used

Can be installed and loaded in 6 lines of code:

```{r, eval=FALSE}
pkgs <- c("devtools", "shiny", "rgdal", "rgeos", "ggmap") # official packages
install.packages(pkgs) 
library(devtools) # enables installation of leaflet
gh_pkgs <- c("rstudio/leaflet", "robinlovelace/stplanr") 
install_github(gh_pkgs) # install packages on github
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T) # load all
```

## RStudio

RStudio Desktop is highly recommended for Shiny development.

```{r, echo=FALSE}
img <- readPNG("/home/robin/repos/robinlovelace.github.io/img/rstudio-shiny-button.png")
grid.raster(img)
```

```{r, echo=FALSE, include=FALSE}
pkgs <- c("devtools", "shiny", "rgdal", "rgeos", "ggmap") # official packages
gh_pkgs <- c("rstudio/leaflet", "robinlovelace/stplanr") 
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T) # load all
```

## Shiny

**shiny** is a framework for creating online interactive data visualisation 'apps'.

> - A framework for making R output **interactive**
> - An 'app' development framework
> - A (small) growing community of developers centred around RStudio
> - A way to structure online tools: separation of GUI and server via `server.R` and `ui.R` files.


## Alternatives

> - **animate** package
> - Google Charts (**googleVis::**) and 
> - **plotly**
> - D3
> - Tableau

For maps

> - Leaflet (supported by Shiny)
> - Google Maps API

## What's shiny good (and not so good) for?

Build's on R's existing strengths

> - Data visualisation (ggplot2)
> - Increasingline widely understood language for querying data
> - Amazing range of add-on packages

*Flexibility*

Not so good for

> - Scalability
> - Database interaction
> - Low-level control 

## What do shiny apps look like?

```{r, echo=FALSE}
shinyAppDir("~/repos/learning-shiny/himod/")
```

## A second example

```{r, echo=FALSE}
shinyAppDir("~/repos/learning-shiny/rentSplit/")
```

## Leaflet

An R interface to the Leaflet JavaScript library, compatible with Shiny.

```{r, eval=FALSE}
cent <- geocode("Girona")
leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(data = cent)
```


# II: Context

## What motivated this work?

> - House of Commons Criticising the DfT for using closed models
> - 'Impact' and 'engagement' increasingly important for research funding
> - Complex data cannot be adequately summarised in a single static graphic
> - We've been contracted by the DfT (£100,000+) to create a 'National Propensity to Cycle Tool' (NPCT)!

See: [cedar.iph.cam.ac.uk/research/modelling/npct-tool](http://www.cedar.iph.cam.ac.uk/research/modelling/npct-tool/)

## Interactive online tools

> - [AURIN](http://aurin.org.au/)
> - Leeds-Bradford cycle path [interactive map](http://www.cyclecityconnect.co.uk/participate.php)
> - DECC's 'Energy 2050' [planning tool](http://2050-calculator-tool.decc.gov.uk/)
> - Walkability index site
> - Interactive download of bicycle paths

# III: Building the National Propensity to Cycle Tool (NPCT)

## Design criteria

- Interactive basemap
- Points, lines and polygons
- Cascading effects of different scenarios
> - Tidy!

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_114.png"))
```


## Early user testing

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_115.png"))
```

## Genius additions: zoom dependent selection; freeze Scope

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_156.png"))
```

## The master app 

```{r, eval=FALSE}
runApp("~/repos/pct-shiny/master/", launch.browser = T)
```

Key components:

```
|-- master
|   |-- master.R
|   |-- server.R
|   |-- ui.R
|   `-- pct-shiny-funs.R 
|-- manchester
|   |-- server.R (links to ../master.R)
|   `-- ui.R (links to ../ui.R)
`-- README.Rmd

```

## Running the app for different cities

```{r, eval=FALSE}
runApp("~/repos/pct-shiny/norwich/", launch.browser = T)
runApp("~/repos/pct-shiny/cambridge/", launch.browser = T)
runApp("~/repos/pct-shiny/manchester/", launch.browser = T)
```

# IV: Sustainable transport planning with R (stplanr)

## Location-dependent loading script

```{r, eval=FALSE}
start_time <- Sys.time() # for timing the script

la <- "manchester" # Name of the local authority
dir.create(paste0("pct-data/", la))

# ... 200 + lines of code here!
# ...

end_time <- Sys.time()
end_time - start_time
```

## Mid-script diagnostic plots

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_157.png"))
```

## Saving routes on the network

```{r, eval=FALSE}
example(gLines2CyclePath)
data(package = "stplanr", "l")
plot(l)
cckey <- Sys.getenv('CYCLESTREET')
routes_fast <- gLines2CyclePath(l)
routes_slow <- gLines2CyclePath(l, "quietest")
lines(routes_fast, col = "red")
lines(routes_slow, col = "green")
```

```{r, echo=FALSE}
data(package = "stplanr", list = list("l", "routes_fast","routes_slow" ))
plot(l)
lines(routes_fast, col = "red")
lines(routes_slow, col = "green")
```

## gLines2CyclePath

```{r}
gLines2CyclePath
```

# V: Esperanzas y sueños

## Un mundo mejor

> - Espero que el Departamento de Transporte utilise nuestra herramienta para prioritisar *donde* y *como* invertir en sistemas de transporte 'post-carbon'
> - Espero que el esfuerzo este util para otros investigadores que quieren un
sistema de transporte mas saludable
>- Sueño que algun dia las decisiones estrategicas se haran usando datos, metodos y software libre
> - Sueño que nuestra herremienta mejorara las decisiones estrategicas en muchos paises del mundo

> - ¿Preguntas?
