---
title: "Testing methods for implementing the 'gender equality' scenario"
author: "Robin Lovelace"
date: "May 3, 2015"
output: word_document
---

The proportion of males vs females cycling is known per zone.
This can be defined as
`pcyclez_m` for the proportion of male cyclists and `pcyclez_f` for the
proportion of female cyclists. Because people are either male or female,
`pcyclez_f = 1 - pcyclez_m`: `pcycle_m + pcycle_f` must equal 1.

Imagine `pcyclez_m = 0.75`
The *ratio* of male to female cyclists is `pcyclez_m / (1 - pcyclez_m)`: 
3:1 in this case.

Now consider that we know the total (`tflow = 50`) and cyclist flow
(`olc = 5`) between two points. Further we know the *ratio* of males vs
females making this trip is 1.5. This can be stated as `ptrips_m = 0.6`:
the proportion of all trips made my males is 0.6 because `1.5 / (1.5 + 1) = 0.6 `.

The question is: what would the total rate of cycling be if women were just as
likely to cycle as men in this set-up?

# Method 1: use ratios assuming `olc_m = olc * pcycle_m`

One approach is to assume that the proportion of males cycling to work for the flow
in question is the same as the proportion of cyclists who are male in the zone; 
that `olc_m / olc = pcycle_m`. This would have the consequence that
`olc_m = 3.75`, the outcome of `5 * 0.75`.

It is fair to assume that the future (gender equal) male:female ratio amongst
cyclists will be the same as the male:female ratio for the flow overall: `1.5:1`.
To ensure that this is the case, the new ratio of female cyclists simply becomes
the current rate divided by this ratio. Clearly `ptrips_f = 1 - ptrips_m`, 0.4,
so 2.5 women are expect to cycle in the future scenario, compared with 1.25 currently.

`slc_gendereq_f = olc_m * ptrips_f / (1 - ptrips_f)`

From here we can calculate the new rate of cycling:
`slc_gendereq = olc_m + slc_gendereq_f`, or 6.25.

To verify this result, let's look at the percent of men and women who cycle.
30 men and 20 women travel along the flow. `3.75 / 30` suggests that 12.5%
of males cycle. `2.5 / 20` is also 12.5%, so the strategy works from the perspective
of the proportion of men and women cycling becoming equal.

# Method 2: assume  `olc_m` scales with the proportion of cyclists

A problem with method 1 is that it assumes that `olc_m = olc * pcycle_m`
regardless of `ptrips_m`. If `ptrips_m` is very low, this become unlikely.
A solution could be to include `ptrips_m` in the estimate of `olc_m`, as
a scaling factor (`pscale`)
which is affected by the ratio of `ptrips_m` multiplied by the proportion
of trips made by bicycle:

```{r, eval=FALSE}
pscale = 1 + (ptrips_m / (1 - ptrips_m) - 1) * (olc / tflow)
```

In this instance, because more males than females make the trip, the dominance
of males in `olc` is set to increase, by a factor of 1.05, by 5%. This results
in an estimate of `olc_m` of 3.9375. This suggests 13.1% of males in the
flow cycle.

```{r, eval=FALSE}
olc_m = olc * pcyclez_m * pscale
```

From this point, the same logic as descibed in Method 1 applies:

```{r, eval=FALSE}
slc_gendereq_f = olc_m * ptrips_f / (1 - ptrips_f)
```

Thus `slc_gendereq` is estimated as 2.625, slightly higher than in Method 1.

# Comparing the two methods

Method 2 has the advantage of including `ptrips_m` in the estimate
of the number of males cycling in the zone. However, it also has the 
disadvantage of complexity. In the absence of a strong reason to prefer
the more complex method the simpler Method 1 is favoured.

Method 2, if verified, could be a useful refinement of Method 1, for use in
future versions of this work.

